Conserving bird populations in the Anthropocene: the significance of non-breeding movements

Version 1

Abstract

Abstract

Background

To conserve bird species threatened by climate change, it is important to understand how environmental factors affected by climate change, such as snow cover, impact their ranges. While this problem is fairly well understood for breeding areas, it remains poorly understood for non-breeding areas. In non-breeding areas, seasonal cycles can strongly influence the distribution of resources during winter. If birds adapt to such changes, they may result in seasonal and directional movement of birds within their non-breeding range. In this case, birds would experience a unique migration pattern - rapid migration between breeding and non-breeding habitats versus a slow migration pattern within their non-breeding range. Their non-breeding range would therefore be dynamic, with potentially important consequences for our understanding of population densities and non-breeding ranges.

Methods

Between 2013-2021, we tracked 43 adult Rough-legged buzzards with solar GPS-GSM loggers. We analyzed their behavior, determined whether the birds showed any directional return migrations during the non-breeding season, and evaluated the differences between the slow migration within their winter range and the quick migration between breeding and non-breeding areas. We also analyzed the vegetation cover of the areas crossed during quick and slow migrations and the role of snow cover in winter migrations.

Results

Our findings revealed that after a quick fall migration through the taiga zone, Rough-legged buzzards continue to migrate during the non-breeding season, albeit at a slower pace across the wooded fields they select as habitat. They avoid complete snow cover and move to escape the progression of the snow cover line from northeast to southwest and back during the winter. As a consequence, Rough-legged buzzards have a dynamic winter range. Thus, the migration pattern of these birds comprises alternating quick and slow phases, resembling the foxtrot dance, which we have named the ‘foxtrot’ migration pattern. Due to this pattern, their winter range displays a dynamic shift of the seasonal center of the population distribution over 1000 km towards the southwest and back throughout the winter.

Conclusions

Our study uncovered a novel bird migration pattern postulated to exist before but poorly understood. This ‘foxtrot migration’ likely occurs in many migratory species inhabiting winter areas with pronounced seasonal cycles. Our findings have implications for conservation efforts in the Anthropocene, where environmental factors such as snow cover can change rapidly and have cascading effects on bird migration. We recommend presenting dynamic winter ranges in species descriptions and range maps so ecologists can use them to develop effective conservation strategies.

Introduction

Global climate change significantly impacts organisms and regions worldwide, with the Arctic experiencing more pronounced effects (Davy & Outten 2020; Previdi, Smith & Polvani 2021). These changes are potentially responsible for the decline in the numbers and ranges of many species and populations (Post et al. 2009; Gilg et al. 2012; Schmidt et al. 2012). Understanding the impact of environmental factors highly susceptible to climate change variation, such as snow cover, is critical for species conservation planning (Lawler 2009; Groves et al. 2012; Reside, Butt & Adams 2018). However, determining the distribution range and the seasonal changes in the local density of highly mobile taxa such as birds can be challenging. Many bird species undergo seasonal migrations, traditionally dividing their range into breeding and non-breeding areas. Most studies of climate change’s impact on bird species have focused on changes in the breeding range, with less emphasis on the non-breeding range (Schmidt et al. 2012). This is primarily because the breeding range is easy to determine from nest counts and does not change during the season. In contrast, the non-breeding range is difficult to define as birds are not tied to any one location, despite spending most of their annual cycle within this range. Moreover, environmental factors affected by climate change can significantly influence the dynamics of the non-breeding range. In recent decades, advanced tracking technology has greatly improved our ability to study bird ecology. We can now follow birds throughout their life cycle and study the influence of environmental factors on the dynamic of their non-breeding range (Wikelski et al. 2007; Flack et al. 2022; Jetz et al. 2022).

The non-breeding range of many bird species can experience dramatic seasonal changes in food availability due to environmental conditions, such as the progressive snow cover line (Bildstein 2006). This line moves – at least in most north-temperate areas exposed to the northwesterly wind systems – from north (-east) to south (-west) and back again in many mid-latitude areas between October and May, affecting food availability for many bird species (Sonerud 1986; Vansteelant, Faveyts & Buckens 2011). We propose that species dependent on this environmental factor move gradually away from the snow cover during the winter and then gradually move in the opposite direction, resulting in a directed and seasonal movement of their winter range. Considering the complete life cycle of such species, they would exhibit a quick migration phase when moving between breeding and non-breeding ranges, followed by a slow migration phase with directed and seasonal movement within the non-breeding range, and finally, another quick migration phase towards the breeding range. Many insectivore bird species wintering in Africa exhibit a similar pattern – relatively fast migration across the Sahara desert and then experiencing gradual directional movements during the non-breeding period (Stach et al. 2012; Jacobsen et al. 2017). Researchers propose that this movement is a consequence of a shift in the insect peak, which is, in turn, influenced by seasonal changes in precipitation (Stach et al. 2012). This type of movement can be compared to the foxtrot dance, which also involves alternating quick and slow phases of movement, both forward and backward. We, therefore, refer to this putative migration as the ‘foxtrot’ migration (Figure 1a) to allow for a quick and easily understandable description. The winter range in species exhibiting this type of migration will then have a dynamic structure that shifts with the ongoing non-breeding migration over time (Figure 1b).

Hypothetical foxtrot migration and dynamic winter range scheme.

a) Foxtrot migration, Q. mig. – Quick migration (spring and fall migrations). b) Dynamic winter range.

This concept of ‘foxtrot’ migration synthesizes migrating and non-breeding movements. Traditionally, studies of bird movement ecology fall into two broad categories: migration studies and studies of movements within breeding and non-breeding ranges (Newton 2008; Berthold, Gwinner & Sonnenschein 2013). Migration studies have focused on the directed seasonal return movements between breeding and non-breeding ranges (Alerstam, Hedenström & Åkesson 2003; Bildstein 2006; Curk et al. 2020). This includes studies of different aspects of migration behavior: reasons for migration (McKinnon et al. 2010; Somveille, Rodrigues & Manica 2018; Pokrovsky et al. 2021), the timing of migration (Lameris et al. 2018; Lehikoinen et al. 2019), duration and length of migration routes (Nilsson et al. 2013), and other aspects. The variability in these characteristics gives rise to a wide range of migratory patterns, which vary greatly in the time taken to migrate. Some species migrate non-stop to their wintering grounds (Battley et al. 2012), some species experience a ‘fly-and-forage’ strategy, where birds hunt during migration and reduce their stopover time (Strandberg and Alerstam 2007), some make long stops to refuel (Shimada et al. 2014). Research in this area has also examined the influence of external environmental factors such as wind, rain, temperature, and day length on bird migration, revealing how these factors can limit food availability and shape migration patterns (Liechti & Bruderer 1998; Alerstam, Hedenström & Åkesson 2003; Curk et al. 2020; Pokrovsky et al. 2021). The second area of research focuses on daily routine movements, dispersal movements, nomadism, and habitat selection within breeding and non-breeding habitats (Greenwood & Harvey 1982; Lenz et al. 2015; Curk et al. 2022). In these studies, it is common to observe long movements in time and distance that are non-directional and non-returning as birds search for food or suitable nesting sites. At the same time, seasonal cycles can significantly affect resource distribution across the wintering range of species. This, in turn, can lead to directional and seasonal bird movements within the wintering range, commonly referred to as migratory behavior. We hypothesize that such a migration pattern and dynamic winter range could be observed in mid-latitude wintering Arctic raptors.

We used the Rough-legged buzzard (Buteo lagopus) as a model species to investigate this phenomenon. The Rough-legged buzzard is an Arctic breeding and mid-latitude wintering raptor (Ferguson-Lees & Christie 2001; Bechard & Swem 2002). Rough-legged buzzards feed mainly on small rodents in the Arctic and its wintering grounds (Tast, Kaikusalo & Lagerström 2010; Pokrovsky et al. 2014). They prefer open areas for hunting, and trees and tall bushes or uplands for resting. In the Arctic, such areas are the southern and typical tundra (Walker et al. 2005), and in the mid-latitudes, areas with fields and patches of forests (wooded fields). The taiga zone, where there is little open space, is unsuitable for them, although they may nest in the northern taiga zone on the border with the tundra zone (Sundell et al. 2004). Snow cover and day length play an important role in their life (Terraube et al. 2015; Curk et al. 2020; Pokrovsky et al. 2021). Rough-legged buzzards can only hunt during the day (Pokrovsky et al. 2021), and heavy snow cover makes hunting for small rodents problematic (Sonerud 1986; Vansteelant, Faveyts & Buckens 2011). These two environmental factors, therefore, affect the availability of prey for Rough-legged buzzards. At the same time, these factors vary considerably in mid-latitudes during winter. Thus, prey availability for Rough-legged buzzards in the mid-latitudes increases until mid-winter if they migrate southwards and after mid-winter if they migrate northwards. We, therefore, assume that Rough-legged buzzards could track prey availability and experience a ‘foxtrot’ migration and dynamic wintering distribution.

For this study, we hypothesized that Rough-legged buzzards would undertake a combined ‘foxtrot’ migration and dynamic winter range in response to seasonal changes in mid-latitude environmental factors. We made the following predictions. 1) Buzzards would exhibit a directional and seasonal movement pattern during the winter, moving from the northeast to the southwest and back again. This would result in a dynamic winter range that would continue to move geographically throughout the season. 2) Winter migrations would differ from fall and spring migrations in duration, extent, speed, and direction. 3) Winter migrations would occur in suitable open habitats, whereas fall and spring migrations would occur in unfitting forested areas. 4) The snow conditions experienced by the Rough-legged buzzards during their winter migration would differ from those in the areas of their winter range where they arrived after their autumn migration, with less snow cover encountered during their winter migration than if they had remained where they arrived at the end of their fall migration.

In the following, we will refer to the spring and fall migrations as the quick migration, the winter migration to the lowest point of latitude as the 1st phase of the slow migration, and the migration from the lowest point of latitude to the starting point of the spring migration as the 2nd phase of the slow migration (Figure 1a).

Material and methods

Dataset

For this study, we tracked 43 adult Rough-legged buzzards (35 females and eight males) with the solar GPS-GSM loggers (e-obs GmbH and UKn – University of Konstanz). The fieldwork was carried out in the Russian Arctic in 2013–2019 at four study sites: Kolguev Island (69°16′N, 48°87′E), Nenetsky Nature Reserve (68°20′N, 53°18′E), Vaigach Island (69°43′N, 60°08′E), and Yamal Peninsula (68°12′N, 68°59′E). For details on capture methods and tag parameters, see Curk et al. (2022); for detailed study descriptions, see Pokrovsky et al. (2015) for Kolguev and Pokrovsky et al. (2019) for Yamal and Nenetsky.

During data pre-processing, we estimated the date of death using the accelerometer or GPS data and removed the tracking data if the bird was dead. We then removed duplicated timestamps and calculated the mean daily positions of each individual. We partitioned the resulting dataset into several periods: 1) breeding, 2) fall migration, 3)1st phase of winter, 4) 2nd phase of winter, and 5) spring migration. We estimated the migration dates – the start and stop dates of the spring and fall migrations – using an iterative search procedure for piecewise regression described by Crawley (2007). We estimated the date between winter’s first and second phases as the day when the mean daily latitude was minimum.

Data analysis

First, we used linear mixed-effects models (R function ‘lmer’ in the library ‘lme4’ (Bates et al. 2015)) to investigate whether or not Rough-legged buzzards migrated during winter. Latitude was the response variable, day of the year was a fixed effect, and individuals and year were included as random effects. Analyses were conducted separately for each migration period (fall, first phase of winter, second phase of winter, and spring). For both phases of the winter migration, we analyzed two additional models with longitude as the response variable instead of latitude. Likelihood ratio tests were used to compare candidate models. The year was not a calendar year but a year between two consecutive breeding seasons. Thus, fall migration, consecutive winter, and consecutive spring have the same value for the year. The day of the year was recalculated consecutively.

Second, we used linear mixed-effects models (R function ‘lmer’ in the library ‘lme4’) to investigate whether migrations’ parameters differ between the migration periods. We analyzed four migration parameters: distance, duration, speed, and direction. The distance was calculated as the distance between two coordinates (start and end of migration) using the R function ’distm’ in the library ‘geosphere’ (Hijmans 2016). The duration was calculated as the number of days between the start and end of migration. Speed was calculated as the ratio of distance to duration. The direction was calculated as the bearing from the start of the migration coordinates to the end of the migration coordinates using the R function ’bearing’ in the library ‘geosphere’ (Hijmans 2016). The migration parameter was used as the response variable, the type of migration as a fixed factor, and individuals as a random factor. Likelihood ratio tests were used to compare candidate models. We considered four different parameters of migration (distance, duration, speed, and direction) and four types of migration (fall, first phase of winter, second phase of winter, and spring). The analysis was done separately for each of the migration parameters. Then, we used post hoc comparisons using the R function ‘emmeans’ in the library ‘emmeans’ (Lenth et al. 2019) to compare the estimated means. In some raptor species, adult females disperse further than males (Mearns & Newton 1984; Serrano et al. 2001; Bildstein 2006; Whitfield et al. 2009). Therefore, we conducted an additional analysis on the effect of sex on migration length using linear mixed-effects models (R function ‘lmer’ in the library ‘lme4’). The migration distance was used as the response variable, sex as a fixed factor, and individuals as a random factor.

Third, we investigated whether vegetation land cover differed between areas crossed during the quick (fall and spring) and slow (winter) migrations. We used the combined Terra and Aqua Moderate Resolution Imaging Spectroradiometer (MODIS) Land Cover Climate Modelling Grid (CMG) (MCD12C1) version 6 dataset (Friedl & Sulla-Menashe 2015). We used a modified Leaf Area Index (LAI) as a classification scheme. We combined all four forest types and savannas into one category (forest) and excluded the categories: water bodies, and unclassified. We, therefore, had five types of vegetation cover: forest, grassland, cropland, shrubland, and urban. We annotated the mean daily positions with the vegetation cover type using the Env-DATA tool (Dodge et al. 2013). We used general linear mixed effects models with a binomial distribution (R function ‘glmer’ in the library ‘lme4’) to investigate whether vegetation cover types differ between migration periods. Presence/absence of the studied vegetation land cover type was used as a response variable, migration type as a fixed factor, and individuals as a random factor. The analysis was done separately for each of the vegetation land cover types.

Fourth, we investigated whether snow cover could drive the slow migration phenomenon.

We then compared the snow cover conditions the birds experienced during the winter with two hypothetical snow cover conditions that the birds would have experienced if they had not migrated during the winter. The first hypothetical snow cover condition would have happened if the birds had stayed where they arrived from the north (i.e., where their fall migration ended). To estimate this parameter, we calculated the winter dynamics of the average snow cover at the minimum convex polygons (MCP) occupied by the birds in October and April (northeast of their winter range). A second hypothetical snow cover condition would be if the birds flew immediately to the southwest and spent the whole winter there. To evaluate this, we calculated the winter dynamics of average snow cover on the MCPs occupied by the birds in January and February (southwest of their winter range). We then compared the values obtained for the real snow cover and two hypothetical snow covers using general linear mixed effects models with a binomial distribution (R function ’glmer’ in the ’lme4’ library). Presence/absence of snow cover was used as a response variable, type of snow cover (real, 1st hypothetical, or 2nd hypothetical) as a fixed factor, and years as a random factor. The analysis was done separately for each month. We then used post hoc comparisons to compare the estimated means, using the R function ‘emmeans’ in the ‘emmeans’ library (Lenth et al. 2019).

We obtained monthly snow cover data with a spatial resolution of ca 500 meters (Global SnowPack MODIS) from the German Aerospace Center (DLR). This product is based on the Moderate Resolution Imaging Spectroradiometer (MODIS) daily snow cover products MOD10A1 and MYD10A1 (version 6 as provided by the National Snow and Ice Data Center NSIDC), which have been processed to remove the gaps due to cloud cover and polar darkness (Dietz, Kuenzer & Dech 2015). These processing steps include a combination of data available from different satellites (Aqua and Terra), 3-day temporal moving window filtering, a regional snow line elevation interpolation relying on a Digital Elevation Model (DEM), and a seasonal filter running through the time series for the whole hydrological year (1st of September through August 31st). The proportion of days in which one pixel is snow-covered per month is referred to here as fractional snow cover and is derived from these daily gap-filled rasters. Five MODIS tiles (h19v03, h20v03, h20v04, h21v03 and h21v04) were mosaicked and re-projected to WGS84. Then, for each month from October to April, we calculated 95% minimum convex polygons (MCPs) for the distribution of Rough-legged buzzards using the R function ‘mcp’ in library ‘adehabitatHR’ (Calenge 2006). We extracted mean snow cover values from each MCP from every monthly snow cover raster separately, using the R library ’raster’ (Hijmans & van Etten 2023).

All calculations were performed using R version 4.2.2 ‘Innocent and Trusting’ (R Development Core Team 2022) and RStudio version 353 ‘Elsbeth Geranium’ (Posit team 2022).

Results

Foxtrot migration – always on the move

Except during the breeding season, Rough-legged buzzard migration continues throughout the year, even after the birds’ arrival at their traditionally recognized ‘wintering grounds’ (Figure 2a, b). For both quick and slow migrations, linear mixed-effects models with the season as a fixed factor received higher support from the likelihood ratio test (p < 0.001, Tables S1-S3). Rough-legged buzzards started their fall migration on 28 September (hereafter mean±sd for the day of the year: 271±11, n=31) and ended on 12 October (285±11, n=33). The mean latitude/longitude where the birds ended their fall migration was 55.57±1.92°/49.35±5.63°. During the winter, birds continued to migrate at a slower pace down to 49.53±2.01° latitude (on 5 February, 36±40, n=23) and 34.29±5.11° longitude (on 24 January, 24±47, n=23). Afterward, during the second phase of the slow winter migration, the birds returned to 55.52±2.63° latitude and 49.79±8.24° longitude to start the spring migration. Rough-legged buzzards started their spring migration to the Arctic on 27 April (117±7, n=27) and arrived at the breeding grounds on 15 May (135±8, n=18).

Quick and slow migration patterns of Rough-legged buzzards throughout an annual cycle, termed ‘foxtrot migration’ for quick referencing.

Q – quick migration. Qf – Quick fall migration (orange), Qs – Quick spring migration (yellow), S1 – Slow migration, 1st phase (light blue), S2 – Slow migration, 2nd phase (dark blue). a) Change in the latitude of 43 Rough-legged buzzards during the year, red line – mean latitude of all birds, black vertical lines – mean dates of start and end of the migration phases, blue vertical line – mean date of the minimum latitude. Grey, sky blue, and piggy pink shaded areas – standard deviation of the means. b) Migration map. c) Difference in the migration parameters between the migration periods. Lines on the direction plot (down, right) represent the mean value for each bird; arrows represent the mean direction.

Quick-slow migration features comparison

During quick migration, individual birds flew greater distances in a shorter time, i.e., at a faster rate, than during slow migration. After arriving at what is traditionally known as the wintering grounds, the direction of migration changed, so the direction of quick and slow migration also differed (Table 1, Figure 2c). The quick migration was 1415±50 km long, whereas the slow migration (one phase) was 1026±55 km, i.e., 389±60 km shorter (p<0.001, Table S4, Figure 2c). During quick migration, birds flew for 15±3 days, and one phase of slow migration lasted 100±4 days, i.e., 85±5 days longer (p<0.001, Table S5, Figure 2c). At the same time, the second phase of slow migration was 54±7 days shorter than the first (p<0.001, Table S5, Figure 2c). The migration speed was 104±6 km/day during the quick phase and 12±7 km/day during the slow phase, i.e., about eight times higher (p<0.001, Table S6, Figure 2c). During the fall migration, birds moved in the SSW direction (7±2 deg), then turned 50±3 deg (p<0.001, Table S7, Figure 2c) to the west and started their 1st slow phase until mid-winter. After that, they turned back to the NEE direction (57±2 deg) and performed their 2nd slow migration phase for several months until they turned 54±3 deg (p<0.001, Table S7, Figure 2c) to the north and started their spring migration. As a result of additional analysis of the effect of sex on migration length, we found no significant difference between the migration distances of males and females (Table S8).

Vegetation land cover during migration

During fast migrations, Rough-legged buzzards cross the forest zone, while during slow migrations, they migrate within the grassland and cropland zone (Figure 3). Rough-legged buzzards migrated fast across the tundra zone on the north in the Arctic and then through the

Vegetation land cover during quick and slow migration.

a) Quick migration. Spring and fall migration together. b) Slow migration. 1st and 2nd phases of the slow migrations together. c) Migration map.

taiga zone. Therefore, during quick migration, the three most common vegetation land cover types were forest (44.5±2.9 %, hereafter, percentage of all mean daily positions during migration annotated with the given vegetation type), shrublands (29.9±3 %), and grasslands (24.7±2.8 %, Figure 3a). During slow migration, the three most common vegetation land cover types were grasslands (65.1±6.2 %), croplands (26.9±6.3 %), and forests (4.9±1.4 %, Figure 3b). According to the linear mixed-effects models, the percentage of all vegetation land cover types differed between the slow and quick migration periods (p<0.001, Table S9), except for the urban lands. Urban lands were more common during the slow than quick migration (Figure 3). However, this type has been annotated for too few birds to make an adequate comparison.

Snow cover – the main reason for the dynamic winter range

During the slow winter migration, Rough-legged buzzards experienced snow cover ranging from 4.8±1.0% in October to 85.2±4.6% in February (Figure 4). If birds spent the winter in the place where they arrived after the fall migration, they would experience snow cover conditions ranging from 4.6±0.6% in October to 99.5±0.1% in February (Figure 4). And if birds fly directly to the southeast and stay there for the whole winter, they would experience snow cover conditions ranging from 1.4±0.2 % in October to 81.1±5.0 % in January (Figure 4). Thus, if birds fly immediately to the southwest and stay there until the end of the winter, they will find conditions with less snow cover in spring (p<0.001, Table S10). And if birds stay where they ended the fall migration, they will find themselves in situations with more snow cover (p<0.001, Table S10). In the latter case, the difference between real and hypothetical situations is not as pronounced (85.2% vs. 99.5%), but more means that snow cover will be close to 100% for several months in this hypothetical situation (Figure 4b).

Slow migration and snow cover conditions.

a) 95% Minimum convex polygons (MCPs) of Rough-legged buzzards during winter. Arrows indicate the direction of the movement across months. OCT – October, NOV – November, DEC – December, JAN – January, FEB – February, MAR – March, APR – April. b) Snow cover conditions for the real situation (black) and two hypothetical situations – if birds spend the winter in the place where they arrived after the fall migration (green) and if birds fly directly to the southwest and stay there all winter (red). Dots represent mean values, error lines – standard errors.

Discussion

Our study has revealed a novel bird migration pattern that likely occurs in many migratory species inhabiting winter areas with pronounced seasonal cycles. This pattern results in the dynamic shift of the birds’ non-breeding range by hundreds of kilometers. In contrast to traditional migration patterns where birds settle in a single location after the initial rapid migration phase in the fall (Bildstein 2006; Newton 2008), the quick-and-slow migration involves a continuous migration at a slower pace through the winter range, followed by a swift return to the breeding site (Figures 2a, b; 4a; 5a). This pattern features a consistent, directional return migration throughout the winter, which is repeated annually, unlike the nomadic movements observed in some bird species during winter (Eng & Schladweiler 1972; Roth & Vetter 2008; Heckscher et al. 2011). We, therefore, termed this migration pattern ‘foxtrot’ migration – a quick movement followed by slow movements in different directions – for quick referencing. Also, unlike the ’fly-and-forage’ migration pattern (Strandberg & Alerstam 2007) and long stopovers during migrations (Shimada et al. 2014) caused by physiological needs, in this case, birds are moving primarily due to external factors that are changing rapidly in the present-day world as a result of climate change. Therefore, this pattern is potentially subject to significant change during the Anthropocene, a change that could have cascading effects on species distributions.

A 2022 Dutch study found a decline in wintering Rough-legged buzzards over the last 40 years (Hornman, Boele & van Winden 2022). On the one hand, this may represent a conservation concern. On the other hand, applying the rationale of the foxtrot migration, the apparent local decline may simply be attributed to climate change, resulting in less comprehensive snow coverage in the northeastern wintering areas of Rough-legged buzzards relative to the Netherlands. Such a shift in snow coverage makes it less probable for the birds to migrate to the Netherlands for overwintering. This proposition is further supported by a study of the winter population dynamics of Rough-legged buzzards in the Netherlands in 2011, showing that the main winter population peak occurred in late December, with many birds migrating (Vansteelant, Faveyts & Buckens 2011). The authors also found that the main migration occurred after heavy snowfall in northern Europe, supporting our ‘foxtrot’ migration hypothesis. We suggest that population changes of highly mobile organisms in many non-breeding areas around the world could be similarly explained by a more or less continuous movement of individuals in response to climate alterations.

Understanding the dynamic range and migration pattern of Rough-legged buzzards is crucial for developing effective conservation strategies for this, but also for other species. In conservation planning, balancing prohibited management activities within the species’ range and the benefits of such activities is a key consideration (Beatley 1992; Jones, Pejchar & Kiesecker 2015; Harfoot et al. 2018; Carrell, Hammill & Edwards 2022). By understanding the dynamic range of the species, a more optimal trade-off can be identified, allowing management activities to be carried out when the species is not present. We recommend that researchers working on species distribution atlases and range maps take this important aspect of a species’ biology into account. In most of the bird atlases and web resources of today (Keller et al. 2020; Audubon 2022), the winter range is usually depicted as a single entity or an area where the species may be present in varying densities (Figure 5b, c), indicating spatial heterogeneity. We propose that temporal heterogeneity be identified as separate zones, indicating periods when the species is abundant, perhaps beyond a threshold value, in the area. To differentiate from spatial heterogeneity, we suggest using lines to define the boundaries of these zones instead of using color shading (Figure 5c) or arrows to show the migration phases (Figure 5a). It is important to note that when mid-winter counts (e.g., Christmas counts) are used to determine the winter range, the spatial heterogeneity observed may be due to temporal heterogeneity. This study indicates that the fact that the majority of birds are far from the northern limit of their winter range in mid-winter does not necessarily mean that they are less abundant there throughout the winter compared to other parts of their winter range.

Winter range representations.

Red – breeding range, blue – winter range. a) Migration cycle of Rough-legged buzzard; arrows represent migration (solid line – quick migration, dashed line – slow migration). b) Winter range without heterogeneity shown. c) Winter range with spatial heterogeneity. d) Winter range with temporal heterogeneity; the numbers indicate the months the birds are common in the outlined area.

While the practical importance of this migratory behavior is clear, understanding the reasons behind this pattern of migration is crucial for predicting potential changes in response to climate change. As in the case of species wintering in subtropical or tropical parts of Africa, the movement patterns are likely driven by the interplay between seasonal cycles and habitat types – similar to what we propose for the alternating periods of fast and slow migration in Rough-legged buzzards. During the quick migration phase, birds pass through habitats that are unfavorable for their survival, while during the slow migration phase, they navigate through more suitable habitats. In our study, the quick migration tracks, as shown in Figures 2 a, b and 3, traverse the taiga zone, which is not conducive to the hunting needs of Rough-legged buzzards, which require open areas for hunting (Ferguson-Lees & Christie 2001; Bechard & Swem 2002). Upon arriving in an appropriate habitat, such as a wooded field and croplands, the birds temporarily halt or slow their migration but then turn from the south to the southwest and gradually migrate in this direction as the winter progresses (Figures 2c and 4). The question arises as to why the birds turn to the southwest and do not continue their migration further south into the steppe zone, which is an open habitat, too. We assume that the lack of forests in the steppe, which serve as crucial roosting sites for Rough-legged buzzards, prevents them from doing so. Our analysis of the locations assigned to forests during the slow migration revealed a scarcity of such sites, with less than 5% of the locations being assigned to forests (Figure 3b).

Despite their scarcity in these habitats, forests appear to hold significant importance for Rough-legged buzzards for nocturnal safety. Another question is why Rough-legged buzzards do not stay in places where their autumn migration has ended. Based on our analysis, we postulate that the primary driver behind the winter migration of Rough-legged buzzards is the southwestward progression of the snow line over the winter season.

The progression of snow cover in the wintering range of Rough-legged buzzards plays a significant role in their winter migration pattern. The winter range of Rough-legged Buzzards experiences a dynamic snow cover with a gradual increase from October to January-February and a subsequent gradual decrease from February to May, moving from the northeastern part of the range to the southwestern part. Correspondingly, the buzzards also move from northeast to southwest during the winter, as if in response to the increasing snow (Figures 2b, 4a). They only partially avoid snow, with snow cover reaching 85% in the southeast of their range by mid-winter. However, they successfully avoid continuous snow cover, with snow cover reaching 100% in the northeast part of their range (Figure 4b). There is a critical limit of snow cover, around 90%, beyond which hunting becomes infeasible. The shortening day length in the northeast compared to the southwest and the fact that Rough-legged buzzards only hunt during the day (Pokrovsky et al. 2021) highlights the significance of even a tiny difference in snow cover. One might wonder why the buzzards would not move to the southwest of the range in the autumn and remain there throughout the winter. This behavior might be attributed to optimizing migration energy costs, with the buzzards moving to the optimal habitat only when the prevailing conditions become unfavorable for their survival. This is in accordance with the theory of optimal migration (Alerstam & Lindström 1990; Houston 1998; Newton 2008). However, some individuals may migrate directly to the southwest, reflecting the plasticity of the migration behavior (Figure S1). The dynamics of snow cover hold a significant influence not only over the winter migration patterns but also in the entire migration cycle of Rough-legged buzzards.

Both quick and slow migration phases in Rough-legged buzzards are influenced by environmental changes that are part of seasonal cycles (Sonerud 1986; Curk et al. 2020). These cycles result in variations in temperature, precipitation, snow cover, and day length, which affect food availability. Latitude affects the extent of these seasonal changes. In autumn, birds in northern breeding grounds face declining temperatures, shorter days, and increased snow cover, prompting a migration to escape these adverse conditions (Alerstam 1991; Houston 1998; Alerstam, Hedenström & Åkesson 2003). However, unfavorable habitats in the southern portion of their breeding range (taiga zone) prevent gradual movement away from adverse environmental conditions. Hence, they perform quick migration to cross the forest zone. Once in favorable habitats, the birds are temporarily insulated from the negative factors they had previously fled, allowing for respite. As the influence of these negative factors eventually reaches the birds’ favorable habitats in the south, they must migrate again, but at a slower pace that corresponds with the gradual approach of these factors. This delay results in a slower migration rate in the first half of winter compared to the second half (Figure 2, Table S2).

In some raptor species, adult females disperse further than males (Mearns & Newton 1984; Serrano et al. 2001; Bildstein 2006; Whitfield et al. 2009). While our analysis does not support this (Table S8), it should be acknowledged that our dataset comprises 35 females, eight males, and no juveniles. The unbalanced sex ratio and absence of immature individuals in our sample may have potential influences towards the observed movement patterns in this study.

The quantification of environmental changes that could prove fatal to bird species presents yet another challenge for conservation efforts in an era of rapid global change. In this study, we have demonstrated the impact of seemingly minor variations in ecological parameters on the species’ behavior. For instance, the presence of small patches of woodland in the winter range might appear crucial to the survival of the Rough-legged buzzard. Elimination of these seemingly minor elements of vegetation cover through management actions could have dire consequences for the species. Similarly, snow cover, while not necessarily a limiting factor in the distribution of a particular species, can become critical at certain thresholds and play a crucial role in shaping species’ behavior. As our study shows, if Rough-legged buzzards fly southwest immediately and stay there for the whole winter, they will experience 25.7% (57.3% vs. 31.6%) less snow cover in March, but this does not force them to adopt this strategy. At the same time, if Rough-legged buzzards stay where they finished their fall migration and don’t move southwest, they will experience 14.4% (99.5% vs. 85.1%) more snow cover in February, and this will force them to migrate. The point is that in this case, although the difference is not as great (14.4% vs. 25.7%), it is the difference between complete snow cover and inconsistent snow cover. This again shows that when we discuss conservation strategies, it is not necessarily important to judge such parameters in terms of numbers or percentages, but rather to understand the context of these parameters in the species’ biology and their importance for the species’ existence. In some contexts, significant environmental changes may not affect the species’ life, but in other contexts, even small changes in the same environmental parameters may be fatal to the species: in short, there may be tipping points in environmental settings that are not predicted by a small linear increase in effect sizes.

Conclusions

The quick-and-slow migration pattern of Rough-legged buzzards (we termed them ‘foxtrot migration’) results in a significant temporal shift of their winter range of over 1000 km. After the birds arrive in their ‘wintering range’, they move slowly but continuously towards the southwest and back during the winter, highlighting the adaptability of birds in response to changing environments. Our findings have crucial implications for conservation efforts in the Anthropocene, where rapidly changing environmental factors can severely impact bird migration (Tucker et al. 2018; Sumasgutner et al. 2021). We suggest that ecologists should include dynamic winter ranges in species descriptions and range maps to develop effective conservation strategies for the bird species. The knowledge about dynamic range shifts is imperative in the face of rapid climate and human-induced habitat changes threatening wildlife and humans.

Supporting information

eLife Assessment

This work presents valuable findings on the non-breeding itinerant behavior of a migratory raptor. With its extensive dataset and solid analytical framework, this work will be of broad interest to researchers investigating the ecological drivers of bird migration. However, the main claim on a novel migration pattern (so-called 'fox-trot migration') is incomplete in light of current knowledge on bird migratory behavior.

Reviewer #1 (Public Review):

This is a paper describing in detail the seasonal movements of a vole-eating raptor, the rough-legged buzzard, from their Arctic breeding areas to the temperate wintering areas and back, in an annual cycle perspective. The basis of the descriptions (using satellite tags) is state of the art, and so are the analyses on aspects of time and space. Of particular relevance is the degree in which this study successfully pinpoints the ecological shaping factors, food availability of course, in this case strongly affected by snow cover (which can be remotely sensed over large areas). The authors claim a new migration pattern called 'foxtrot' with phases with rapid and phases with slower migration movements.

My concern with this paper is the framing. A story on the how and why of these continental movements in response to snow and other habitat features would be a grand contribution.

Reviewer #2 (Public Review):

This preprint by Pokrovsky and coworkers is a descriptive study reporting on non-breeding itinerant behaviour of an intrapalearctic migratory raptor, the rough-legged buzzard, and relating such non-breeding movements to snow cover across the European non-breeding range. The article is based on long-term GPS tracking data from a relatively large sample of individuals (n=43) that were equipped with state-of-the-art tracking devices in the Russian Arctic during 2013-2019. The results show that, upon breeding, buzzards migrated rapidly to southern non-breeding areas, located in open areas north of the Black and Caspian seas, where they perform continuous directional movements at a slower pace, initially moving SW (Oct to Jan) and then progressively moving NE (Feb to Apr) before embarking on rapid spring migration. It is suggested that such itinerant behaviour follows variation (expansion and retreat) of snow cover across the non-breeding range.

The results are definitely useful for researchers investigating the ecological drivers of bird movement patterns. The paper is generally well-written and the analytical framework is solid. However, there are significant weaknesses in the theoretical framework, unwarranted claiming of novelty, and interpretation of the data. Below are key points that the authors may wish to consider.

1. The authors underemphasize the fact that what they term 'fox-trot' migration is actually a well-known pattern for many other migratory species, both in the Nearctic and in the Afro-Palearctic migration systems. Such behaviour has previously been identified as 'itinerant', involving an alternation of stopovers and movements between different short-term non-breeding residency areas, and it seems that the pattern the authors report for this particular species is perfectly in line with such previous evidence. For instance, this is well-documented among migratory raptors, such as the Montagu's harrier, a lesser kestrel or black kite, that exploit Sahelian savannahs, where large spatio-temporal variation in greenness and hence resource availability occurs. And, besides the mentioned cuckoos and nightingales, there are studies of red-backed shrikes suggesting the same, as well as of tree swallows in the Nearctic. Therefore, the authors should avoid claiming novelty for this study and introducing unnecessary and confusing new terms in the literature (i.e. the 'fox-trot' migration patterns) when these are definitely not strictly needed as they have been previously observed and defined otherwise. Reference to all this previous body of literature is only hinted at and should be considerably expanded. The final sentence of the abstract, involving a general recommendation for future work, is definitely not warranted. Sentences such as 'We used the rough-legged buzzard as a model..." are also similarly unwarranted. This is simply a descriptive study reporting on such behaviour in yet another migratory species. The predictions paragraph is also overlong and could be considerably condensed.

2. The term 'migration' associated to so-called 'fox-trot' movements (see Fig. 1) is also highly confusing and possibly incorrect, as it is not in line with the commonly accepted definition of 'migration' (i.e. mass back and forth movements from the same areas). Apparently, the authors do not provide any evidence that the birds are moving back and forth from the same areas during the non-breeding period (i.e., there is no mention of site fidelity between early and late wintering areas, but judging from fall and spring migration distances it seems this is definitely not the case). 'Non-breeding itinerancy' is clearly a more appropriate term to describe this behaviour. More generally, the reference to 'winter migration', which is often mentioned in the manuscript, is not correct and should be amended.

3. The current title is unnecessarily general (it may recall rather a review or meta-analysis) and not adequately describing the content of the manuscript. It is not at all clear how the terms 'Conservation' and 'Anthropocene' are related to the content of the study (unless one believes that this is because any study of wildlife is aimed at its conservation, which is of course untrue, and that the study has been performed in the Anthropocene, which is the case for all wildlife studies carried out after 1950-1960). In order to be informative, the title should more tightly reflect the content of the article. A valid alternative would be 'Itinerant non-breeding behaviour of an intra-Palaearctic migratory raptor', far more adequate and informative. Although it might be worthwhile mentioning the association between movements and snow cover (or ecological conditions more generally) already in the title, perhaps that link is too indirect as currently reported in the manuscript. There are several possibilities to provide a more direct link between movements and snow cover, such as e.g. performing habitat selection analysis with respect to snow cover. Plotting temporal progression of snow cover (average) against movements (e.g. by showing monthly home ranges against snow cover) would help visualizing the association between snow cover and movement patterns.

4. The text, particularly the Introduction and (even more so) the Discussion, would benefit from profound reframing in light of the above comments. Any link to conservation is too weak and should be removed or considerably toned down. Moreover, the species is not of conservation interest (IUCN = Least Concern), as it has an extremely large range and population size, with largely fluctuating and non-declining populations (whose dynamics are related to Arctic small rodent cycles). Unless the authors are able to make prediction on how these movements will be affected by climate change (e.g. by using species distribution models or similar approaches), the link to the Anthropocene and to conservation is mostly unwarranted. In general, reference to 'winter' should be avoided and replaced with 'non-breeding season', which is a more general term.

Author response:

The following is the authors’ response to the current reviews.

We sincerely appreciate the reviewer’s dedication to evaluating our manuscript and raising essential considerations regarding the classification of the migration behavior we described. While the reviewer suggests that this behavior aligns with the concept of itinerancy, we contend that it represents a distinct phenomenon, albeit with similarities, as both involve the non-breeding movements of birds. We acknowledge that our manuscript did not adequately address this distinction and have considered the reviewer’s feedback. In our response, we clarify the difference between the described phenomenon and itinerancy. Our revised manuscript will include a new section in the Discussion to address this issue comprehensively.

In the first part of the review, the reviewer emphasizes that the pattern we are describing is consistent with itinerancy. Regardless of the terminology used, we want to highlight the existence of two different types of migratory behavior, both of which involve movement in non-breeding areas.

The first type, called itinerancy, was first described by Moreau in 1972 in “The Palaearctic-African Bird Migration Systems.” As noted by the reviewer, this behavior involves an alternation of stopovers and movements between different short-term non-breeding residency areas. They usually occur in response to food scarcity in one part of the non-breeding range, causing birds to move to another part of the same range. These movements typically cover distances of 10 to 100 kilometers but are neither continuous nor directional. Moreau (1972) defined itinerancy as prolonged stopovers, normally lasting several months, primarily in tropical regions. He noted observations of certain species disappearing from his study areas in sub-Saharan Africa in December and others appearing, suggesting they may have multiple home ranges during the non-breeding season. Subsequent research, as mentioned by the reviewer, has confirmed itinerancy in many species, particularly among Palaearctic-African migrants in sub-Saharan Africa. In particular, the Montagu’s Harrier has been extensively studied in this regard. The reviewer rightly points out that our study does not include recent findings on this species. In our revised version, we will include references to recent studies, such as those by Trierweiler et al. (2013, Journal of Animal Ecology, 82:107-120) and Schlaich et al. (2023, Ardea, 111:321-342), which show that Montagu’s Harrier has an average of 3-4 home ranges separated by approximately 200 kilometers. These studies suggest that the species spends approximately 1.5 months at each site, with the most extended period typically observed at the last site before migrating to the breeding grounds.

In the second type, birds undertake a post-breeding migration, arrive in their non-breeding range, and then gradually move in a particular direction throughout the season. This continuous directional movement covers considerable distances and continues throughout the non-breeding period. In our study, this movement covered about 1000 km, comparable to the total migration distance of Rough-legged Buzzards of about 1500 km. As observed in our research, these movements are influenced by external factors such as snow cover. In such cases, the progression of snow cover in a south-westerly direction during winter can prevent birds from finding food, forcing them to continue migrating in the same direction. In essence, this movement represents a prolonged phase of the migration process but at a slower pace. Similar behavior has been documented in buzzards, as reported by Strandberg et al. (2009, Ibis 151:200-206). Although several transmitters in their study stopped working in mid-winter, the authors observed a phenomenon they termed ‘prolonged autumn migration.’

In the second part of the review, the reviewer questions the need to distinguish between the two behaviors we have discussed. However, we believe these behaviors differ in their structure (with the first being intermittent and often non-directional, whereas the second is continuous and directional) and in their causes (with the first being driven by seasonal food resource cycles and the second by advancing snow cover). We therefore argue that it is worth distinguishing between them. To differentiate these forms of non-breeding movement, we propose to use ‘itinerancy’ for the first type, as described initially by Moreau in 1972, and introduce a separate term for the second behavior. Although ‘slow directional itinerancy’ could be considered, we find it too cumbersome.

Moreover, ‘itinerancy’ in the literature refers not only to non-breeding movements but also to the use of different nesting sites, e.g., Lislevand et al. (2020, Journal of Avian Biology: e02595), reinforcing its association with movements between multiple sites within habitats. We, therefore, propose that the second behavior be given a distinct name. We acknowledge the reviewer’s point that we did not adequately address this distinction in the Discussion and plan to include a separate section in our paper’s revised version. In the third part of his review, the reviewer suggests an alternative title. Another reviewer, Dr Theunis Piersma, suggested the current title during the first round of reviewing, and we have chosen his version.

In the fourth part of the review, the reviewer questions whether it is appropriate to discuss the conservation aspect of this study. This type of non-breeding movement raises concerns about accurately determining non-breeding ranges and population dynamics for species that exhibit this behavior. We believe that accurate determination of range and population dynamics is critical to conservation efforts. While this may be less important for species breeding in Europe and migrating to Africa, for which monitoring breeding territories is more feasible, it’s essential for Arctic and sub-Arctic breeding species. Large-scale surveys in these regions have historically been challenging and have become even more so with the end of Arctic cooperation following Russia’s war with Ukraine (Koivurova, Shibata, 2023). For North America and Europe, non-breeding abundance is typically estimated once per season in mid-winter. In North America, these are the so-called Christmas counts (which take place once at the end of December), and in Europe, they are the IWC counts mentioned by the reviewer (as follows from their official website - “The IWC requires a single count at each site, which should be repeated each year. The exact dates vary slightly from region to region, but take place in January or February”). Because of such a single count in mid-winter, non-breeding habitats occupied in autumn and spring will be listed as ‘uncommon’ at best, while south-western habitats where birds are only present in mid-winter will be listed as ‘common.’ However, the situation will be reversed if we consider the time birds spend in these habitats.

The reviewer also highlights the introduction’s unconventional structure and information redundancy at the beginning. We have chosen this structure and provided basic explanations to improve readability for a wider audience, given eLife’s readership. At the same time, we will certainly take the reviewers’ feedback into account in the revised version. We plan to include the references to modern itinerancy research mentioned above and to add a section on itinerancy to the Discussion.

We appreciate the reviewer’s input and sincerely thank them for their time and effort in reviewing our paper. While we may not fully agree on the classification of the behavior we describe, we value the opportunity to engage in discussion and believe that presenting arguments and counterarguments to the reader is beneficial to scientific progress.

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations For The Authors):

I much enjoyed reading this manuscript, that is, once I understood what it is about. Titles like "Conserving bird populations in the Anthropocene: the significance of non-breeding movements" are a claim to so-called relevance, they have NOTHING to do with the content of the paper, so once I understood that this paper was about the "Quick quick slow: the foxtrot migration of rough-legged buzzards is a response to habitat and snow" (an alternative title), it was becoming very interesting. So the start of the abstract as well as the introduction is very tedious, as clearly much trouble is taken here to establish reputability. In my eyes this is unnecessary: eLife should be interested in publishing such a wonderful description of such a wonderful migrant in a study that comes to grips with limiting factors on a continental scale!

We sincerely appreciate your time and effort in reviewing our manuscript. Thank you for your appreciation of our study.

We agree that the focus of the article should be changed from conservation to migration patterns. We have rewritten the Introduction and Discussion as suggested. We have added the application of this pattern including conservation at the end of the Discussion by completely changing Figure 5. We have also changed the title to the suggested one.

Not sure that the first paragraph statements that seek to downplay what we know about wintering vs breeding areas are valid (although I see what purpose they serve). Migratory shorebirds have extensively been studied in the nonbreeding areas, for example, including movement aspects (see, as just one example, Verhoeven, M.A., Loonstra, A.H.J., McBride, A.D., Both, C., Senner, N.R. & Piersma, T. (2020) Migration route, stopping sites, and non breeding destinations of adult Black tailed Godwits breeding in southwest Fryslân, The Netherlands. Journal of Ornithology 162, 61-76) and there are very impressive studies on the winter biology of migrants across large scale (for example in Zwarts' Living on the Edge book on the Sahel wetlands). Think also about geese and swans and about seabirds!

We have rewritten the first paragraph and it now talks about patterns of migratory behavior. We have also rewritten the second paragraph, now it is devoted to studies of movements in the non-breeding period. We explain how our pattern differs from those already studied and give references to the papers you mentioned.

Directional movements in nonbreeding areas as a function of food (in this case locusts) have really beautifully been described by Almut Schlaich et al in JAnimEcol for Montagu's harriers.

We have added Montagu's harrier example in the second paragraph of the Introduction and the Discussion. We have added a reference to Schlaich and to Garcia and Arroyo, who suggested that Montagu's harriers have long directional migrations during the non-breeding period.

Once the paper starts talking buzzards, and the analyses of the wonderful data, all is fine. It is a very competent analysis with a description of a cool pattern.

Thank you for your appreciation of our study. We hope the revised version is better and clearer.

However, i would say that it is all a question of spatial scale. The buzzards here respond to changes in food availability, but there is not an animal that doesn't. The question is how far they have to move for an adequate response: in some birds movements of 100s of meters may be enough, and then anything to the scale of rough-legged buzzards.

In the new version of the manuscript, we emphasize that this is a large distance (about 1000 km), comparable to the distance of the fall and spring migrations (about 1400 km) in lines 70-72 of the Introduction and 379-383 of the Discussion.

And actually, several of the shorebirds I know best also do a foxtrot, such as red knots and bar-tailed godwits moulting in the Wadden Sea, then spending a few months in the UK estuaries, before returning to the Wadden Sea before the long migrations to Arctic breeding grounds. The publication of the rough-legged buzzard story may help researchers to summarize patterns such as this too. Mu problem with this paper is the framing. A story on the how and why of these continental movements in response to snow and other habitat features would be a grand contribution. Drop Anthropocene, and rethink whether foxtrot should be introduced as a hypothesis or a summary of cool descriptions. I prefer the latter, and recommend eLife to go with that too, rather than encourage "disconnected frames that seek 'respectability'" Good luck, theunis piersma

We thank the reviewer again for his valuable comments and suggestions. We have changed the framing to the suggested one and removed the Anthropocene from the article.

Reviewer #2 (Recommendations For The Authors):

We sincerely appreciate the time and effort you have taken to review our manuscript. We have carefully considered all of your comments, including both public and author comments, and provided detailed responses to each of them below. In addition, we would like to address the most important public comments.

We agree with the suggestion to shift the focus of the article from conservation to migration patterns. Accordingly, we have rewritten both the Introduction and Discussion sections to focus on migration behavior rather than conservation.

However, we respectfully disagree with the suggestion that the migration patterns we describe are synonymous with itinerancy. We acknowledge that our original presentation may have been unclear and may have hindered full understanding. In the revised version, we provide a detailed analysis of migratory behavior in the Introduction that describes how our pattern differs from itinerancy. We also revisit this distinction in the Discussion section. We have also carefully revised Figure 1 to improve clarity and avoid potential misunderstandings.

Regarding the applicability of the described migration pattern, we acknowledge that the Rough-legged Buzzard is not listed as an endangered species. However, we believe that our findings have practical implications. We have moved our discussion of this issue to the end of the Discussion section and have completely revised Figure 5. While the overall population of Rough-legged Buzzards is not declining, certain regions within its range are experiencing declines. We show that this decline does not warrant listing the species as endangered. Instead, it may represent a redistribution within the non-breeding range - a shift in range dynamics. We use the example of the Rough-legged Buzzard to illustrate this concept and emphasize the importance of considering such dynamics when assessing the conservation status of species in the future.

We also acknowledge that the hypothesis of this form of behavior has been proposed previously for Montagu's Harrier, and we have included this information in the revised manuscript. In addition, we agree that the focus on the Anthropocene is unnecessary in this context and have therefore removed it.

We believe that these revisions significantly improve the clarity and robustness of the manuscript, and we are grateful for your insightful comments and suggestions.

As a general comment, please note that including line numbers (as it is the standard in any manuscript submission) would facilitate reviewers providing more detailed comments on the text.

We apologize for this oversight and have added line numbers to our revised manuscript.

Dataset: unclear what is the frequency of GPS transmissions. Furthermore, information on relative tag mass for the tracked individuals should be reported.

We have included this information in our manuscript (L 157-163). We also refer to the study in which this dataset was first used and described in detail (L 164).

Data pre-processing: more details are needed here. What data have been removed if the bird died? The entire track of the individual? Only the data classified in the last section of the track? The section also reports on an 'iterative procedure' for annotating tracks, which is only vaguely described. A piecewise regression is mentioned, but no details are provided, not even on what is the dependent variable (I assume it should be latitude?).

Regarding the deaths. We only removed the data when the bird was already dead. We have corrected the text to make this clear (L 170).

Regarding the iterative procedure. We have added a detailed description on lines 175-188.

Data analysis: several potential issues here:

(1) Unclear why sex was not included in all mixed models. I think it should be included.

Our dataset contains 35 females and eight males. This ratio does not allow us to include sex in all models and adequately assess the influence of this factor. At the same time, because adult females disperse farther than males in some raptor species, we conducted a separate analysis of the dependence of migration distance on sex (Table S8) and found no evidence for this in our species. We have written a separate paragraph about this. This paragraph can be found on lines 356-360 of the new manuscript.

(2) Unclear what is the rationale of describing habitat use during migration; is it only to show that it is a largely unsuitable habitat for the species? But is a formal analysis required then? Wouldn't be enough to simply describe this?

Habitat use and snow cover determine the two main phases (quick and slow) of the pattern we describe. We believe that habitat analysis is appropriate in this case and that a simple description would be uninformative and would not support our conclusions.

(3) Analysis of snow cover: such a 'what if' analysis is fine but it seems to be a rather indirect assessment of the effect of snow cover on movement patterns. Can a more direct test be envisaged relating e.g. daily movement patterns to concomitant snow cover? This should be rather straightforward. The effectiveness of this method rests on among-year differences in snow cover and timing of snowfall. A further possibility would be to demonstrate habitat selection within the entire non-breeding home range of an individual in relation snow cover. Such an analysis would imply associating presence-absence of snow to every location within the non-breeding range and testing whether the proportion of locations with snow is lower than the proportion of snow of random locations within the entire non-breeding home range (95% KDE) for every individual (e.g. by setting a 1/10 ratio presence to random locations).

The proposed analysis will provide an opportunity to assess whether the Rough-legged Buzzard selects areas with the lowest snow cover, but will not provide an opportunity to follow the dynamics and will therefore give a misleading overall picture. This is especially true in the spring months. In March-April, Rough-legged Buzzards move northeast and are in an area that is not the most open to snow. At this time, areas to the southwest are more open to snow (this can be seen in Figure 4b). If we perform the proposed analysis, the control points for this period would be both to the north (where there is more snow) and to the south (where there is less snow) from the real locations, and the result would be that there is no difference in snow cover.

A step-selection analysis could be used, as we did in our previous work (Curk et al 2020 Sci Rep) with the same Rough-legged Buzzard (but during migration, not winter). But this would only give us a qualitative idea, not a quantitative one - that Rough-legged Buzzards move from snow (in the fall) and follow snowmelt progression (in the spring).

At the same time, our analysis gives a complete picture of snow cover dynamics in different parts of the non-breeding range. This allows us to see that if Rough-legged Buzzards remained at their fall migration endpoint without moving southwest, they would encounter 14.4% more snow cover (99.5% vs. 85.1%). Although this difference may seem small (14.4%), it holds significance for rodent-hunting birds, distinguishing between complete and patchy snow cover. Simultaneously, if Rough-legged Buzzards immediately flew to the southwest and stayed there throughout winter, they would experience 25.7% less snow cover (57.3% vs. 31.6%). Despite a greater difference than in the first case, it doesn't compel them to adopt this strategy, as it represents the difference between various degrees of landscape openness from snow cover.

We write about this in the new manuscript on lines 385-394.

Results: it is unclear whether the reported dispersion measures are SDs or SEs. Please provide details.

For the date and coordinates of the start and end of the different phases of migration, we specified the mean, sd, and sample size. We wrote this in line 277. For the values of the parameters of the different phases of the migration (duration, distance, speed, and direction), we used the mean, the standard error of the mean, and the confidence interval (obtained using the ‘emmeans’ package). We have indicated this in lines 302-303 and the caption of Table 1 (L 315) and Figure 2 (L 293-294). For the values of habitat and snow cover experienced by the Rough-legged Buzzards, we used the mean and the error of the mean. We reported this on lines 322 and 337 and in Figures 3 (L 332-333) and 4 (L 355-356).

Discussion: in general, it should be reshaped taking into account the comments. It is overlong, speculative and quite naive in several passages. Entire sections can be safely removed (I think it can be reduced by half without any loss of information). I provide some examples of the issues I have spotted below. For instance, the entire paragraph starting with 'Understanding....' is not clear to me. What do you mean by 'prohibited management' options? Without examples, this seems a rather general text, based on unclear premises when related to the specific of this study. Some statements are vague, derive from unsubstantiated claims, and unclear. E.g. "Despite their scarcity in these habitats, forests appear to hold significant importance for Rough-legged buzzards for nocturnal safety". I could not find any day-night analysis showing that they actually roost in forests during nighttime. Being a tundra species, it may well be possible that rough-legged buzzards perceive forests as very dangerous habitats and that they prefer instead to roost in open habitats. Analysing habitat use during day and night during the non-breeding period may be of help to clarify this. Furthermore, considering the fast migration periods, what is the flight speed during day and night above forests? Do these birds also migrate at night or do they roost during the night? Perhaps a figure visualizing day and night track segments could be of help (or an analysis of day vs. night flight speed) (there are several R packages to annotate tracks in relation to day and night). This is an example of another problematic statement: "The progression of snow cover in the wintering range of Rough-legged buzzards plays a significant role in their winter migration pattern." The manuscript does not contain any clear demonstration of this, as I wrote in my previous comments. Without such evidence, you must considerably tone down such assertions. But since providing a direct link is certainly possible, I think that additional analyses would clearly strengthen your take-home message.

The paragraph starting with "The quantification of environmental changes that could prove fatal to bird species presents yet another challenge for conservation efforts in an era of rapid global change." is quite odd. Take the following statement "For instance, the presence of small patches of woodland in the winter range might appear crucial to the survival of the Rough-legged buzzard. Elimination of these seemingly minor elements of vegetation cover through management actions could have dire consequences for the species.". It is based on the assumption that minor vegetation elements play a key role in the ecology of the species, without any evidence supporting this. Does it have any sense? I could safely say exactly the opposite and I would believe it might even be more substantiated.

We agree with these comments.

We have completely rewritten this section. As suggested, we have shortened it by removing statements that were not supported by the research. We have completely removed the statements about "prohibited management". We have also removed the statement that "forests appear to be of significant importance to Rough-legged buzzards for nocturnal safety" and everything associated with that statement, e.g. the statement about "small elements of vegetation cover", etc. We do believe that this statement is true in substance, but we also agree that it is not supported by the results and requires separate analysis. At the same time, we believe that this is a topic for a separate study and would be redundant here. Therefore, we leave it for a separate publication.

Conclusion paragraph: I believe this severely overstates the conservation importance of this study. That the results have "crucial implications for conservation efforts in the Anthropocene, where rapidly changing environmental factors can severely impact bird migration" seems completely untenable to me. What is the evidence for such crucial implications? For instance, these results may suggest that climate change, because global warming is predicted to reduce snow cover in the non-breeding areas, might well be beneficial for populations of this species, by reducing non-breeding energy expenditure and improving non-breeding survival. I think statements like these are simply not necessary, and that the study should be more focused on the actual results and evidence provided.

We have completely rewritten this section. We removed the reference to the Anthropocene and focused on migratory behavior and migration patterns.

Version 2

Abstract

Abstract

Advancements in tracking technologies have revolutionized our understanding of bird migrations, revealing a diverse array of migratory behaviors. We propose a new pattern of migratory behavior termed ‘foxtrot migration,’ characterized by alternating quick and slow phases during the non-breeding period. This behavior involves directional and continuous movements, distinct from traditional seasonal itinerancy or forage-on-fly migration. Using the Rough-legged buzzard as a model species, we confirmed the presence of foxtrot migration and ‘dynamic non-breeding ranges’, driven by environmental factors such as snow cover dynamics. We advocate for accurate representation of dynamic ranges on maps and emphasize the need to consider range dynamics when assessing species conservation status. Our findings underscore the importance of understanding complex migratory behaviors in the face of environmental change, facilitated by advancements in tracking technologies. This knowledge is crucial for effective conservation strategies amid ongoing global environmental challenges.

Introduction

In recent decades, advanced tracking technology has greatly improved our knowledge of bird migrations (Wikelski et al. 2007; Jetz et al. 2022; Kays & Wikelski 2023). In the 19th century, the finding of a White stork (Ciconia ciconia) in Germany with an African arrow led to the discovery that birds did not hibernate at the bottom of water bodies during the winter but rather made extensive seasonal migrations (Richter & Bick 2018). Subsequent studies of bird migration in the 20th century using ringing techniques revealed a diversity of seasonal migratory behaviors, including 1) propensity to migrate (migrants, residents, partial migrants), 2) differences among populations (leapfrog, chain, parallel migrations), 3) connectivity and philopatry (constancy of breeding and non-breeding locations), 4) length of migration routes (long-distance and short-distance migrants), 5) variations in flight paces during migration (with stopovers, forage-on-fly, nonstop), and 6) tendencies to move within a single season (breeding and non-breeding itinerancy) (Alerstam, Hedenström & Åkesson 2003; Newton 2008; Alerstam 2011; Berthold, Gwinner & Sonnenschein 2013). Different combinations of these behaviors contribute to a wide variety of bird migration patterns (Chapman et al. 2014; Alerstam & Bäckman 2018; Lislevand et al. 2020). However, the advent of 21st-century technology, such as GPS transmitters, provides a higher resolution of movement and the ability to observe more details of migratory behavior (Wikelski et al. 2007; Jetz et al. 2022). Just as advances in image resolution allow us to see finer details in images, improvements in bird tracking technology now allow us to identify previously undiscovered migratory behaviors, revealing new patterns.

Current technologies allow observing forms of migratory behavior that are intermediate or integrative to those described above. Itinerancy was initially defined as the movements of Afro-Palearctic migrants during the non-breeding season in the Sahel, with a combination of stopovers and flights in between (Moreau 1972). Subsequently, researchers investigated this behavior for multiple species wintering in Africa, attributing their movements during the non-breeding season to changes in food availability predicted by NDVI (Zwarts et al. 2012; Schlaich et al. 2016; Thorup et al. 2017; Verhoeven et al. 2021). In 1998, Garcia and Arroyo hypothesized that certain species, like the Montagu’s harrier (Circus pygargus), might not only cover tens to hundreds of kilometers between various sites in their winter range but could also demonstrate directional, continuous movement throughout the entire non-breeding season. This movement may span a distance similar to their migration distance (ca. 1000 km), essentially indicating a prolonged, slower-paced continuation of migration (García & Arroyo 1998). Because this behavior is intermediate between seasonal movement (itinerancy) and the actual migration process, we can describe it as either ‘persistent directional non-breeding itinerancy’ or ‘very slow forage-on-fly migration in the non-breeding area’. Both options are clumsy and can confuse. We suggest naming this behavior with a new term to avoid confusion with established terms such as ‘itinerancy’ for any seasonal movement and ‘forage-on-fly’ denoting movement from breeding to non-breeding sites (Alerstam, Hake & Kjellén 2006; Alerstam 2011; Lopez-Ricaurte et al. 2024). Considering this behavior as an extension of migration and looking at the complete life cycle of such species, it includes a quick phase during the transition between breeding and non-breeding areas, followed by a slow phase of directed and seasonal movement within the non-breeding area, and finally another quick phase towards the breeding area. Analogous to the alternating fast and slow movements of the foxtrot dance, we propose the term ‘foxtrot migration’ for this seasonal movement to provide a concise and easily understandable description (Figure 1a). Consequently, we suggest to term the non-breeding range of species exhibiting this migration type as ‘dynamic range’ (Figure 1b).

Foxtrot migration and dynamic winter range scheme.

a) Foxtrot migration. The color bar along the y-axis corresponds to the color coding of the habitats in panel b of the chart on the right. b) Dynamic winter range.

The concept of dynamic range, derived from foxtrot migration, is critical to understanding the current range declines and shifts observed in many species. Typically, a species’ range is divided into areas of common and uncommon, with temporal heterogeneity limited to breeding and non-breeding distinctions (Keller et al. 2020; Billerman et al. 2024). However, suppose a species exhibits a foxtrot migration, moving gradually in one direction and back again during the non-breeding season. In this case, it becomes inaccurate to represent its non-breeding range as static. This discrepancy is particularly pronounced when ranges are mapped using mid-winter surveys, such as Christmas counts. During this period, the species may be present at the outer limits of its non-breeding range, leading to a false classification as common in that area when, in fact, it is rare if the entire non-breeding season is considered. Similarly, extended negative population dynamics of a species in a particular part of its non-breeding range could be misinterpreted as a threat to the entire population. Indeed, if this species has a foxtrot migration, range dynamics may have changed, possibly resulting in the species ceasing to migrate to that particular part of its range. Advanced technologies now make it possible the identification of foxtrot migration and range dynamics. This, in turn, facilitates accurate range mapping and proper assessment of threats to the species.

This pattern of behavior has been proposed in species that overwinter in Africa, where non-breeding movements are primarily related to the distribution of insect abundance (García & Arroyo 1998). At the same time, it is reasonable to assume that this behavior may also be characteristic of species that spend the non-breeding season in mid-latitudes, where food availability may be linked to snow cover dynamics. These dynamics have changed particularly rapidly in recent decades due to global warming (Davy & Outten 2020; Previdi, Smith & Polvani 2021). As a result, global climate change may significantly affect birds that breed in the Arctic and winter in mid-latitudes. The Arctic is experiencing more pronounced effects of global change than other regions, with declines in the abundance and range of several species and populations (Post et al. 2009; Gilg et al. 2012; Schmidt et al. 2012). Therefore, understanding how snow cover dynamics affect Arctic species across habitats throughout the annual cycle is essential today. Furthermore, elucidating the foxtrot migration and dynamic non-breeding range of mid-latitude wintering Arctic birds may provide valuable information on expected future changes in their non-breeding range.

We used the Rough-legged buzzard (Buteo lagopus) as a model species to investigate this phenomenon. The Rough-legged buzzard is an Arctic breeding and mid-latitude wintering raptor (Ferguson-Lees & Christie 2001; Bechard & Swem 2002). Rough-legged buzzards feed mainly on small rodents in the Arctic and its wintering grounds (Tast, Kaikusalo & Lagerström 2010; Pokrovsky et al. 2014). They prefer open areas for hunting, and trees and tall bushes or uplands for resting. In the Arctic, such areas are the southern and typical tundra (Walker et al. 2005), and in the mid-latitudes, areas with fields and patches of forests (wooded fields). The taiga zone, where there is little open space, is unsuitable for them, although they may nest in the northern taiga zone on the border with the tundra zone (Sundell et al. 2004). Snow cover and day length play an important role in their life (Terraube et al. 2015; Curk et al. 2020; Pokrovsky et al. 2021). Rough-legged buzzards can only hunt during the day (Pokrovsky et al. 2021), and heavy snow cover makes hunting for small rodents problematic (Sonerud 1986; Vansteelant, Faveyts & Buckens 2011). These two environmental factors, therefore, affect the availability of prey for Rough-legged buzzards. At the same time, these factors vary considerably in mid-latitudes during winter. Thus, prey availability for Rough-legged buzzards in the mid-latitudes increases until mid-winter if they migrate southwards and after mid-winter if they migrate northwards. We, therefore, assume that Rough-legged buzzards could track prey availability and experience a foxtrot migration and dynamic non-breeding range.

For this study, we hypothesized that Rough-legged buzzards undertake a combined foxtrot migration and dynamic non-breeding range in response to seasonal changes in mid-latitude environmental factors. We made the following predictions. 1) Buzzards would exhibit a directional and seasonal movement pattern during the non-breeding period, moving from the northeast to the southwest and back again. This would result in a dynamic non-breeding range that would continue to move geographically throughout the season. 2) Non-breeding movements would occur in suitable open habitats, whereas fall and spring migrations would occur in unfitting forested areas. 3) Non-breeding movements would differ from fall and spring migrations in duration, extent, speed, and direction, however would continue throughout the season. 4) During non-breeding movements, Rough-legged buzzards will experience less snow cover than if they had stayed where they arrived at the end of the fall migration.

In the following, we will refer to the spring and fall migrations as the quick phase, the non-breeding movement to the lowest point of latitude as the 1st part of the slow phase, and the movement from the lowest point of latitude to the starting point of the spring migration as the 2nd part of the slow phase (Figure 1).

Material and methods

Dataset

For this study, we tracked 43 adult Rough-legged buzzards (35 females and eight males) with the solar GPS-GSM loggers (e-obs GmbH and UKn – University of Konstanz). The fieldwork was carried out in the Russian Arctic in 2013-2019 at four study sites: Kolguev Island (69°16′N, 48°87′E), Nenetsky Nature Reserve (68°20′N, 53°18′E), Vaigach Island (69°43′N, 60°08′E), and Yamal Peninsula (68°12′N, 68°59′E). For details on capture methods and tag parameters, see Curk et al. (2022); for detailed study descriptions, see Pokrovsky et al. (2015) for Kolguev and Pokrovsky et al. (2019) for Yamal and Nenetsky.

During data pre-processing, we estimated the date of death using the accelerometer or GPS data and removed the tracking data if the bird was dead. We then removed duplicated timestamps and calculated the mean daily positions of each individual. We partitioned the resulting dataset into several periods: 1) breeding, 2) fall migration, 3)1st part of winter, 4) 2nd part of winter, and 5) spring migration. We estimated the migration dates – the start and stop dates of the spring and fall migrations – using an iterative search procedure for piecewise regression described by Crawley (2007). We estimated the date between winter’s first and second parts as the day when the mean daily latitude was minimum.

Data analysis

First, we used linear mixed-effects models (R function ‘lmer’ in the library ‘lme4’ (Bates et al. 2015)) to investigate whether or not Rough-legged buzzards migrated during winter. Latitude was the response variable, day of the year was a fixed effect, and individuals and year were included as random effects. Analyses were conducted separately for each migration period (fall, first phase of winter, second phase of winter, and spring). For both phases of the winter migration, we analyzed two additional models with longitude as the response variable instead of latitude. Likelihood ratio tests were used to compare candidate models. The year was not a calendar year but a year between two consecutive breeding seasons. Thus, fall migration, consecutive winter, and consecutive spring have the same value for the year. The day of the year was recalculated consecutively.

Second, we used linear mixed-effects models (R function ‘lmer’ in the library ‘lme4’) to investigate whether migrations’ parameters differ between the migration periods. We analyzed four migration parameters: distance, duration, speed, and direction. The distance was calculated as the distance between two coordinates (start and end of migration) using the R function ‘distm’ in the library ‘geosphere’ (Hijmans 2016). The duration was calculated as the number of days between the start and end of migration. Speed was calculated as the ratio of distance to duration. The direction was calculated as the bearing from the start of the migration coordinates to the end of the migration coordinates using the R function ‘bearing’ in the library ‘geosphere’ (Hijmans 2016). The migration parameter was used as the response variable, the type of migration as a fixed factor, and individuals as a random factor. Likelihood ratio tests were used to compare candidate models. We considered four different parameters of migration (distance, duration, speed, and direction) and four types of migration (fall, first phase of winter, second phase of winter, and spring). The analysis was done separately for each of the migration parameters. Then, we used post hoc comparisons using the R function ‘emmeans’ in the library ‘emmeans’ (Lenth et al. 2019) to compare the estimated means. In some raptor species, adult females disperse further than males (Mearns & Newton 1984; Serrano et al. 2001; Bildstein 2006; Whitfield et al. 2009). Therefore, we conducted an additional analysis on the effect of sex on migration length using linear mixed-effects models (R function ‘lmer’ in the library ‘lme4’). The migration distance was used as the response variable, sex as a fixed factor, and individuals as a random factor.

Third, we investigated whether vegetation land cover differed between areas crossed during the quick (fall and spring) and slow (winter) migrations. We used the combined Terra and Aqua Moderate Resolution Imaging Spectroradiometer (MODIS) Land Cover Climate Modelling Grid (CMG) (MCD12C1) version 6 dataset (Friedl & Sulla-Menashe 2015). We used a modified Leaf Area Index (LAI) as a classification scheme. We combined all four forest types and savannas into one category (forest) and excluded the categories: water bodies, and unclassified. We, therefore, had five types of vegetation cover: forest, grassland, cropland, shrubland, and urban. We annotated the mean daily positions with the vegetation cover type using the Env-DATA tool (Dodge et al. 2013). We used general linear mixed effects models with a binomial distribution (R function ‘glmer’ in the library ‘lme4’) to investigate whether vegetation cover types differ between migration periods. Presence/absence of the studied vegetation land cover type was used as a response variable, migration type as a fixed factor, and individuals as a random factor. The analysis was done separately for each of the vegetation land cover types.

Fourth, we investigated whether snow cover could drive the slow migration phenomenon. We then compared the snow cover conditions the birds experienced during the winter with two hypothetical snow cover conditions that the birds would have experienced if they had not migrated during the winter. The first hypothetical snow cover condition would have happened if the birds had stayed where they arrived from the north (i.e., where their fall migration ended). To estimate this parameter, we calculated the winter dynamics of the average snow cover at the minimum convex polygons (MCP) occupied by the birds in October and April (northeast of their winter range). A second hypothetical snow cover condition would be if the birds flew immediately to the southwest and spent the whole winter there. To evaluate this, we calculated the winter dynamics of average snow cover on the MCPs occupied by the birds in January and February (southwest of their winter range). We then compared the values obtained for the real snow cover and two hypothetical snow covers using general linear mixed effects models with a binomial distribution (R function ‘glmer’ in the ‘lme4’ library). Presence/absence of snow cover was used as a response variable, type of snow cover (real, 1st hypothetical, or 2nd hypothetical) as a fixed factor, and years as a random factor. The analysis was done separately for each month. We then used post hoc comparisons to compare the estimated means, using the R function ‘emmeans’ in the ‘emmeans’ library (Lenth et al. 2019).

We obtained monthly snow cover data with a spatial resolution of ca 500 meters (Global SnowPack MODIS) from the German Aerospace Center (DLR). This product is based on the Moderate Resolution Imaging Spectroradiometer (MODIS) daily snow cover products MOD10A1 and MYD10A1 (version 6 as provided by the National Snow and Ice Data Center NSIDC), which have been processed to remove the gaps due to cloud cover and polar darkness (Dietz, Kuenzer & Dech 2015). These processing steps include a combination of data available from different satellites (Aqua and Terra), 3-day temporal moving window filtering, a regional snow line elevation interpolation relying on a Digital Elevation Model (DEM), and a seasonal filter running through the time series for the whole hydrological year (1st of September through August 31st). The proportion of days in which one pixel is snow-covered per month is referred to here as fractional snow cover and is derived from these daily gap-filled rasters. Five MODIS tiles (h19v03, h20v03, h20v04, h21v03 and h21v04) were mosaicked and re-projected to WGS84. Then, for each month from October to April, we calculated 95% minimum convex polygons (MCPs) for the distribution of Rough-legged buzzards using the R function ‘mcp’ in library ‘adehabitatHR’ (Calenge 2006). We extracted mean snow cover values from each MCP from every monthly snow cover raster separately, using the R library ‘raster’ (Hijmans & van Etten 2023).

All calculations were performed using R version 4.2.2 ‘Innocent and Trusting’ (R Development Core Team 2022) and RStudio version 353 ‘Elsbeth Geranium’ (Posit team 2022).

Results

Foxtrot migration – always on the move

Except during the breeding season, Rough-legged buzzard migration continues throughout the year, even after the bird’ arrival at their traditionally recognized ‘wintering ground’ (Figure 2a, b). For both quick and slow phases of the foxtrot migration, linear mixed-effects models with the season as a fixed factor received higher support from the likelihood ratio test (p < 0.001, Tables S1-S3). Rough-legged buzzards started their fall migration (quick phase) on 28 September (hereafter mean±sd for the day of the year: 271±11, n=31) and ended on 12 October (285±11, n=33). The mean latitude/longitude where the birds ended their fall migration was 55.57±1.92°/49.35±5.63° (Figure 2a). During the winter, birds continued to migrate at a slower pace down to 49.53±2.01° latitude (on 5 February, 36±40, n=23) and 34.29±5.11° longitude (on 24 January, 24±47, n=23). Afterward, during the second part of the slow phase, the birds returned to 55.52±2.63° latitude and 49.79±8.24° longitude to start the spring migration (Figure 2a). Rough-legged buzzards started their spring migration to the Arctic on 27 April (117±7, n=27) and arrived at the breeding grounds on 15 May (135±8, n=18).

Foxtrot migration of Rough-legged buzzards.

Q – quick phase. Qf – Quick fall phase (orange), Qs – Quick spring phase (yellow), S1 – Slow phase, 1st part (light blue), S2 – Slow phase, 2nd part (dark blue). a) Change in the latitude of 43 Rough-legged buzzards during the year, red line – mean latitude of all birds, black vertical lines – mean dates of start and end of the migration phases, blue vertical line – mean date of the minimum latitude. Grey, sky blue, and piggy pink shaded areas – standard deviation of the means. b) Migration map. c) Difference in the migration parameters between the migration phases. Lines on the direction plot (down, right) represent the mean value for each bird; arrows represent the mean direction.

Quick-slow phase features comparison

During quick phase, individual birds flew greater distances in a shorter time, i.e., at a faster rate, than during slow phase. After arriving at what is traditionally known as the wintering grounds, the direction of migration changed, so the direction of quick and slow phases also differed (Table 1, Figure 2c). The quick phase was 1415±50 km long, whereas the slow phase (one part) was 1026±55 km, i.e., 389±60 km shorter (p<0.001, Table S4, Figure 2c). During quick phase, birds flew for 15±3 days, and one part of slow phase lasted 100±4 days, i.e., 85±5 days longer (p<0.001, Table S5, Figure 2c). At the same time, the second part of slow phase was 54±7 days shorter than the first (p<0.001, Table S5, Figure 2c). The migration speed was 104±6 km/day during the quick phase and 12±7 km/day during the slow phase, i.e., about eight times higher (p<0.001, Table S6, Figure 2c). During the fall migration, birds moved in the SSW direction (7±2 deg), then turned 50±3 deg (p<0.001, Table S7, Figure 2c) to the west and started their 1st slow phase until mid-winter. After that, they turned back to the NEE direction (57±2 deg) and performed their 2nd slow phase for several months until they turned 54±3 deg (p<0.001, Table S7, Figure 2c) to the north and started their spring migration. As a result of additional analysis of the effect of sex on migration length, we found no significant difference between the migration distances of males and females (Table S8).

Vegetation land cover during migration

During quick phases, Rough-legged buzzards cross the forest zone, while during slow phase, they migrate within the grassland and cropland zone (Figure 3). Rough-legged buzzards migrated fast across the tundra zone on the north in the Arctic and then through the taiga zone. Therefore, during quick phase, the three most common vegetation land cover types were forest (44.5±2.9 %, hereafter, percentage of all mean daily positions annotated with the given vegetation type), shrublands (29.9±3 %), and grasslands (24.7±2.8 %, Figure 3a). During slow phase, the three most common vegetation land cover types were grasslands (65.1±6.2 %), croplands (26.9±6.3 %), and forests (4.9±1.4 %, Figure 3b). According to the linear mixed-effects models, the percentage of all vegetation land cover types differed between the slow and quick phases (p<0.001, Table S9), except for the urban lands. Urban lands were more common during the slow than quick phase (Figure 3). However, this type has been annotated for too few birds to make an adequate comparison.

Vegetation land cover during quick and slow phases.

a) Quick phase (spring and fall periods together). b) Slow phase (1st and 2nd parts together). c) Migration map.

Snow cover – the main reason for the dynamic winter range

During the slow phase of the foxtrot migration, Rough-legged buzzards experienced snow cover ranging from 4.8±1.0% in October to 85.2±4.6% in February (Figure 4). If birds spent the winter in the place where they arrived after the fall migration, they would experience snow cover conditions ranging from 4.6±0.6% in October to 99.5±0.1% in February (Figure 4). And if birds fly directly to the southeast and stay there for the whole winter, they would experience snow cover conditions ranging from 1.4±0.2 % in October to 81.1±5.0 % in January (Figure 4). Thus, if birds fly immediately to the southwest and stay there until the end of the winter, they will find conditions with less snow cover in spring (p<0.001, Table S10). And if birds stay where they ended the fall migration, they will find themselves in situations with more snow cover (p<0.001, Table S10). In the latter case, the difference between real and hypothetical situations is not as pronounced (85.2% vs. 99.5%), but more means that snow cover will be close to 100% for several months in this hypothetical situation (Figure 4b).

Snow cover conditions during the slow phase of the foxtrot migration.

a) 95% Minimum convex polygons (MCPs) of Rough-legged buzzards during winter. Arrows indicate the direction of the movement across months. OCT – October, NOV – November, DEC – December, JAN – January, FEB – February, MAR – March, APR – April. b) Snow cover conditions for the real situation (black) and two hypothetical situations – if birds spend the winter in the place where they arrived after the fall migration (green) and if birds fly directly to the southwest and stay there all winter (red). Dots represent mean values, error lines – standard errors.

In some raptor species, adult females disperse further than males (Mearns & Newton 1984; Serrano et al. 2001; Bildstein 2006; Whitfield et al. 2009). While our analysis does not support this (Table S8), it should be acknowledged that our dataset comprises 35 females, eight males, and no juveniles. The unbalanced sex ratio and absence of immature individuals in our sample may have potential influences towards the observed movement patterns in this study.

Discussion

Our study confirmed the existence of a previously hypothesized bird migration pattern (García & Arroyo 1998) characterized by alternating quick and slow phases (Figure 1). Following the breeding season, birds quickly traverse unfavorable habitats and proceed slowly through winter in favorable environments influenced by external factors. Subsequently, they quickly cross unfavorable terrain once more to reach the breeding region in spring (Figure 2). This pattern, termed ‘foxtrot migration’, distinguishes itself from ‘itinerancy’ by featuring directional and continuous movements. Additionally, it differs from ‘forage-on-fly migration’ as it occurs during the non-breeding period. We propose referring to the non-breeding area as ‘dynamic area’, reflecting its temporal heterogeneity, where the region with the highest bird density continually changes throughout the non-breeding period.

Our study affirmed the presence of foxtrot migration and a dynamic non-breeding range in Rough-legged Buzzards. The taiga zone posed as unfavorable habitat during the quick phase of foxtrot migration (Figure 3), given the difficulty for Rough-legged Buzzards to locate open areas for hunting. Conversely, the grassland and cropland zone served as favorable habitats during the slow phase throughout the entire non-breeding period (Figure 3), offering numerous open areas for hunting. Snow cover was the external factor driving their continual 1000 km southwest movement during winter (Figure 4). Our analysis revealed that if Rough-legged Buzzards remained at their fall migration endpoint without moving southwest, they would encounter 14.4% more snow cover (99.5% vs. 85.1%, Figure 4). Although this difference may seem small (14.4%), it holds significance for rodent-hunting birds, distinguishing between complete and patchy snow cover. Simultaneously, if Rough-legged Buzzards immediately flew to the southwest and stayed there throughout winter, they would experience 25.7% less snow cover (57.3% vs. 31.6%, Figure 4). Despite a greater difference than in the first case, it doesn’t compel them to adopt this strategy, as it represents the difference between various degrees of landscape openness from snow cover. This indicates ecological tipping points in the environment that are not predicted by a slight linear increase in effect size.

This pattern is likely to be observed in many migratory species with distinct seasonal cycles in their non-breeding range. In our study, we focused on an Arctic species that resides in the mid-latitudes during the non-breeding period. In general, foxtrot migration is likely to be a common feature among species where snow cover significantly influences food availability, especially those specializing in rodents. Originally proposed for birds in Africa, where seasonal changes regulate food availability, it’s likely that many insectivorous birds that spend the non-breeding season in Africa also exhibit this form of migration. This type of migration and dynamic non-breeding range is widespread both geographically and taxonomically, with obvious conservation implications.

The conservation implications

The conservation implications of our study are twofold: 1) the use of mid-winter (Christmas) bird surveys to determine non-breeding range may yield inaccurate results for species with foxtrot migrations (Figure 5a), and 2) declines in abundance within a particular segment of the non-breeding range may indicate changes in range dynamics rather than widespread declines in species abundance (Figure. 5b).

The use of foxtrot migration research for the assessment of the non-breeding range and conservation status of a particular species.

a) Christmas counts, often used to determine winter range, may give a misleading representation of non-breeding range for species with foxtrot migration. b) Changing the conservation status of a species based on reduced abundance in a particular area may not be appropriate for species with foxtrot migrations. Red – breeding range, blue – winter range. Latin numbers indicate the month the birds are present in the area.

Christmas Bird Counts provide insight into the mid-winter distribution of a species. For example, based on our study of the Rough-legged Buzzard, the species is predominantly present in the southwestern portion of its non-breeding range during this period, with only a small proportion present in the rest of the region. As a result, a map of the non-breeding range may inaccurately depict the species as common in the southwest and extremely rare in the northeast. This map would be inaccurate because, during the entire non-breeding period, Rough-legged Buzzards spend both fall and spring in the northeastern part and only mid-winter in the southwesternmost part.

To address this, continuous year-round GPS tracking of the species provides a means to track bird locations throughout the non-breeding season, facilitating the creation of accurate distribution maps, particularly for species that exhibit foxtrot migrations and dynamic ranges. We advocate representing temporal heterogeneity (range dynamics) on maps as distinct zones, denoting periods when the species is abundant in a given area. To distinguish temporal from spatial heterogeneity, we recommend using lines to delineate the boundaries of these zones, rather than color shading, and incorporating numbers to denote the months of species abundance in a given zone (Figure 5a). We suggest ecologists include dynamic non-breeding ranges in descriptions and range maps for foxtrot migratory bird species.

Population counts for a species are often limited to a portion of its range. Therefore, conclusions about conservation status drawn from such counts may be misleading. A decline in abundance within a particular portion of the non-breeding range may indicate changes in range dynamics rather than a general decline in the species. For example, climate change may affect snow cover dynamics, reducing its intensity in northern regions. As a result, species whose range dynamics depend on snow cover may choose to remain in the northern areas and not migrate as far south as they traditionally have. Despite this shift in range dynamics, overall species abundance may remain unchanged. Similar patterns have affected Rough-legged Buzzards in some areas of the European non-breeding range.

A 2022 Dutch study found a decline in wintering Rough-legged buzzards over the last 40 years (Hornman, Boele & van Winden 2022). On the one hand, this may represent a conservation concern. On the other hand, applying the rationale of the foxtrot migration, the apparent local decline may simply be attributed to climate change, resulting in less comprehensive snow coverage in the northeastern wintering areas of Rough-legged buzzards relative to the Netherlands. Such a shift in snow coverage makes it less probable for the birds to migrate to the Netherlands for overwintering. This proposition is further supported by a study of the winter population dynamics of Rough-legged buzzards in the Netherlands in 2011, showing that the main winter population peak occurred in late December, with many birds migrating (Vansteelant, Faveyts & Buckens 2011). The authors also found that the main migration occurred after heavy snowfall in northern Europe, supporting our foxtrot migration explanation for this decline. Therefore, investigating the dynamic range and foxtrot migration is critical to understanding a species’ range and effectively assessing its conservation status.

Significant environmental changes, including changes in snowpack dynamics, have occurred in recent decades that may have significant impacts on bird migration (Tucker et al. 2018; Sumasgutner et al. 2021). In order to understand the effects of global environmental change on bird migration behavior, there is an urgent need for in-depth investigations into the intricate relationship between migration patterns and evolving environmental factors. The advent of new technologies in recent decades has expanded our understanding of bird migration, enabling the identification of novel behavioral patterns and facilitating the study of their susceptibility to various environmental influences. This knowledge is becoming increasingly urgent in the face of rapid climate change and anthropogenic habitat alteration, which pose threats to both wildlife and human populations.

Conclusions

Our study has identified and characterized a new pattern of migratory behavior, the ‘foxtrot migration’, along with the associated concept of ‘dynamic range’. This discovery has significant implications for conservation strategies and adequate representation of non-breeding habitats. As animal tracking technology advances, we expect to discover more new details about animal migrations and, consequently, new patterns of migratory behavior. With these technological developments, we are gaining a clearer understanding of the complex patterns and behaviors that drive animal migrations, providing valuable insights for conservation efforts and deepening our understanding of the natural world.

Supporting information

eLife Assessment

This work presents a useful finding on non-breeding itinerant behavior of a migratory raptor. With its extensive dataset and analytical framework, this work is of interest to researchers investigating the ecological drivers of bird migration. However, the main claim on a novel migration pattern (so-called 'fox-trot migration') is incomplete in light of current knowledge on bird migratory behavior.

Reviewer #2 (Public Review):

This preprint by Pokrovsky and coworkers is a descriptive study reporting on non-breeding itinerant behaviour of an intrapalearctic migratory raptor, the rough-legged buzzard, and relating such non-breeding movements to snow cover across the European non-breeding range. The article is based on long-term GPS tracking data from a relatively large (n=43) sample of individuals that were equipped with state-of-the-art tracking devices in the Russian Arctic during 2013-2019. The results show that, upon breeding, buzzards migrated rapidly to southern non-breeding areas, located in open areas north of the Black and Caspian seas, where they perform continuous directional movements at a slower pace, initially moving SW (Oct to Jan) and then progressively moving NE (Feb to Apr) before embarking on rapid spring migration. It is suggested that such itinerant behaviour follows variation (expansion and retreat) of snow cover across the non-breeding range.

The results are potentially useful for researchers investigating the ecological drivers of bird movement patterns. The analytical framework appears solid, although some details on the analyses (requested during the previous review round) are still unclear and have not been modified despite explicit requests. Significant weaknesses in the theoretical framework persist in the revised version, including unwarranted claiming of novelty, overselling of importance of the study, and overinterpretation of the data. Below are key points that the authors did not consider when revising their manuscript.

(1) The authors underemphasize the fact that what they term 'fox-trot' migration is actually a well-known patterns for many other migratory species, both in the Nearctic and in the Afro-Palearctic migration systems. Such behaviour has previously been identified as 'itinerant' or 'non-breeding itinerancy', involving an alternation of stopovers and movements between different short-term non-breeding residency areas, or even slow continuous movements, and it seems that the pattern the authors report for this particular species is perfectly in line with such previous evidence. For instance, this is well-documented among migratory raptors, such as the Montagu's harrier, lesser kestrels or black kites, that exploit Sahelian savannahs, where large spatio-temporal variation in greenness and hence resource availability occurs. And, besides the mentioned cuckoos and nightingales, there are studies of red-backed shrikes suggesting the same, as well as of tree swallows in the Nearctic. Therefore, the authors should avoid claiming novelty for this study and introducing unnecessary and confusing new terms in the literature (i.e. the 'fox-trot' migration patterns) when these are definitely not strictly needed as they have been previously observed and defined otherwise. Sentences such as 'We used the rough-legged buzzard as a model..." are also similarly unwarranted. This is simply a descriptive studies reporting on such behaviour in yet another migratory species. The whole introduction is pervaded by a faulty logic. The authors first introduce a new (unwarranted) term based on previous evidence from other studies (none of which felt any need to introduce and describe it); then they assume, for unclear reasons, that the species they are studying should behave in that way; even more worryingly, based on these assumptions, they make specific predictions on how this species should behave, without any sound biological reason for these predictions. I admit I hardly see any scientific logic in this approach.

(2) The species has a very standard migration for a short-distance migrant, by all means. It moves to non-breeding areas, and once there it slowly moves towards the south in autumn and back again in spring, until it departs for pre-breeding migration. This is no different from other trans-Saharan migratory raptor species that spend the non-breeding period in the Sahel. Whether the species perform any short/medium term stopover (frequenting the same are for some days) during the non-breeding stage (between end of autumn migration and onset of spring migration), as is the case of most species showing non-breeding itinerancy, is not reported. The authors only show a slower pace of movement during the non-breeding period compared to autumn and spring migration, without providing any further details. This hinders comparisons with other previous studies.

(3) The current title is unnecessarily general (it may recall rather a review or meta-analysis) and not adequately describing the content of the manuscript. In order to be informative, the title should more tightly reflect the content of the article. A valid alternative would be 'Itinerant non-breeding behaviour of an intra-Palaearctic migratory raptor', as it would be far more adequate and informative.

(4) The text, particularly the Introduction and the Discussion, would greatly benefit from profound reframing in light of the above comments. Upon reading the first sentence of the introduction, it looks surprising that the authors based their suggestion for 'fox-trot' migration based on a very outdated article on the migration of Montagu's harrier based on sparse ring recovery data which merely suggests the existence of 'movements' within the non-breeding areas (i.e. non-breeding itinerancy), while subsequent large scale satellite tracking studies of this species provided compelling evidence for non-breeding area itinerancy (and again, no mention of 'fox-trot' whatsoever). The discussion is entirely framed around potential issues related to accurate monitoring of population size and trends, which the author surprisingly refer to 'conservation implications'. As I already mentioned in my previous review, the 'conservation implications' of this study are nearly negligible. At best, it suggests that caution should be applied when interpreting population trends of migratory species based on non-breeding area counts only, a pattern that is already well known for decades (consider the long-running IWC coordinated by Wetlands International!). In addition, Christmas Bird Count, a long-term monitoring program of AOS, is mentioned without any accurate reference to what it actually is, assuming that any reader would be familiar with a very peculiar monitoring scheme of the Nearctic region.

The final paragraph epitomizes how authors are overstating the importance of this study, claiming for non-existent novelty and even 'discovery': "Our study has identified and characterized a new pattern of migratory behavior, the 'foxtrot migration', along with the associated concept of 'dynamic range'. This discovery has significant implications for conservation strategies and adequate representation of non-breeding habitats".

Author response:

The following is the authors’ response to the current reviews.

Reviewer #4

We sincerely appreciate the time and effort you have taken to review our manuscript. We followed your recommendations to polish the text and make it easier to understand.

Regarding terms and terminology, we changed “non-breeding” everywhere in the text to “over- wintering.”

Regarding the title, as it was suggested by reviewer #1 as his recommendation, we tried to find a compromise and make the changes you suggested but left part of the suggestion from reviewer #1. So, now it’s “Foxtrot migration and dynamic over-wintering range of an arctic raptor”

Thank you for highlighting the importance of snow cover and changes in snow cover as a possible factor of over-wintering movements. We appreciate your feedback and have explored several approaches to address this issue. Specifically, we examined how both snow cover extent and changes in snow cover influenced movement distance. However, we found no effect of either factor on movement distance.

Our data show that birds leave their sites in October and move southwest, even though snow cover is minimal at that time. They also leave their sites in November and in subsequent months, regardless of the snow cover levels. Thus, we observed no pattern of birds leaving sites when snow cover reaches a specific threshold (e.g., 75-80%). Similarly, we found no evidence of birds staying in areas with a certain snow cover extent (e.g., 30%), nor did they leave sites when snow cover increased by a specific amount (e.g., by 10 or 20%).

It is possible that more experienced birds anticipate that October plots will become inaccessible later in the winter and, therefore, leave early without waiting for significant snow accumulation. Alternatively, other factors, such as brief heavy snowfalls, may trigger movement, even if these do not lead to sustained increases in snow cover. Multiple factors, possibly acting asynchronously, could also play a role. This complexity adds an interesting dimension to the study of ecological patterns. However, in this study, we chose to focus on describing the migration pattern itself and its impact on aspects like over-winter range determination and population dynamics. While we have prioritized this approach, we remain committed to further analyzing the data to uncover additional details about this behavior.

In response to your suggestion, we have expanded the Methods sections to clarify that we tested the effects of snow cover and changes in snow cover on distance (Lines 241-246); the Results section (Lines 348-349). We have also included the relevant plots in the Supplementary Materials. In the Discussion, we noted that this approach did not reveal any significant dependence and acknowledged that this issue requires further investigation (Lines 422-459).

---------

The following is the authors’ response to the previous reviews.

Reviewer #2:

We sincerely appreciate the time and effort you have taken to review our manuscript.

First of all, we apologize for publishing the preprint without incorporating certain adjustments outlined in our earlier response, particularly in the Methods section. This was due to an oversight regarding the different versions of the manuscript. We have corrected this mistake. Our response to the feedback on this section (Methods), with line numbers of the changes made, is immediately below this response. In addition, we have included the units of measurement (mean and standard deviation) in both the results and figure captions for clarity.

To focus on the main point regarding wintering strategies, we acknowledge that in the previous versions, this aspect was inadequately addressed and caused some confusion. In the revised edition, both the Introduction and the Discussion have been thoroughly reworked.

As you suggested, we have removed the long introductory paragraph and all references to foxtrot migrations from the Introduction. As a result, the Introduction is now short and to the point. In the second paragraph, we explain why we propose the wintering strategies outlined (L74-81).

In the Discussion, we've added a substantial new section at the beginning that discusses different wintering strategies. We have also updated Figure 4 accordingly. Previously, we erroneously suggested that Montagu's harrier and other African-Palaearctic migrants might adopt wintering strategies similar to those we describe. Upon further investigation, however, we found that almost all African-Palaearctic migrants exhibit an itinerant wintering strategy. Conversely, the strategy we describe is primarily observed in mid-latitude wintering species.

We have shown that, unlike itinerancy, the birds in our study don't pause for 1-2 months at multiple non-breeding sites, but instead migrate significant distances, up to 1000 km, throughout the winter. Furthermore, unlike itinerancy, the sites they reach are consistently snow-free throughout the year. Following the logic of publications on Montagu's harriers (Schlaich et al. 2023), our birds do not wait for favorable conditions at the next site, as is typical of itinerancy. Moreover, this behavior is influenced by external factors such as snow cover dynamics and occurs primarily in mid-latitudes. Researchers studying a species similar to our subject, the Common buzzard, observed a similar pattern and termed it "prolonged autumn migration" rather than itinerancy. Although their transmitters stopped working in mid-winter, precluding a full observation of the annual cycle, they captured the essence of continued migration at a slower pace, distinct from itinerancy. We've detailed all of these findings in a new section.

In addition, we acknowledge the mischaracterization of the implications of our research as ‘Conservation implications’ and have corrected this to ‘Mapping ranges and assessing population trends’, as you suggested.

Finally, we've rewritten the Conclusion, removing overly grandiose statements and simply summarizing the main findings.

We appreciate your time and effort in reviewing our manuscript. With your invaluable input, it has become clearer, more concise, and easier to understand.

Dataset: unclear what is the frequency of GPS transmissions. Furthermore, information on relative tag mass for the tracked individuals should be reported.

We have included this information in our manuscript (L 115-122). We also refer to the study in which this dataset was first used and described in detail (L 123).

Data pre-processing: more details are needed here. What data have been removed if the bird died? The entire track of the individual? Only the data classified in the last section of the track? The section also reports on an 'iterative procedure' for annotating tracks, which is only vaguely described. A piecewise regression is mentioned, but no details are provided, not even on what is the dependent variable (I assume it should be latitude?).

Regarding the deaths, we only removed the data when the bird was already dead. We estimated the date of death and excluded tracking data corresponding to the period after the bird's death. We have corrected the text to make this clear (L 130-131).

Regarding the piecewise regression. We have added a detailed description on lines 136-148.

Data analysis: several potential issues here:

(1) Unclear why sex was not included in all mixed models. I think it should be included.

Our dataset contains 35 females and eight males (L116). This ratio does not allow us to include sex in all models and adequately assess the influence of this factor. At the same time, because adult females disperse farther than males in some raptor species, we conducted a separate analysis of the dependence of migration distance on sex (Table S8) and found no evidence for this in our species. We have written about that in the Methods (L177-181) and after in the Results (L277-278).

(2) Unclear what is the rationale of describing habitat use during migration; is it only to show that it is a largely unsuitable habitat for the species? But is a formal analysis required then? Wouldn't be enough to simply describe this?

Habitat use and snow cover determine the two main phases (quick and slow) of the pattern we describe. We believe that habitat analysis is appropriate in this case, and a simple description would be uninformative and not support our conclusions.

(3) Analysis of snow cover: such a 'what if' analysis is fine but it seems to be a rather indirect assessment of the effect of snow cover on movement patterns. Can a more direct test be envisaged relating e.g. daily movement patterns to concomitant snow cover? This should be rather straightforward. The effectiveness of this method rests on among-year differences in snow cover and timing of snowfall. A further possibility would be to demonstrate habitat selection within the entire non-breeding home range of an individual in relation snow cover. Such an analysis would imply associating presenceabsence of snow to every location within the non-breeding range and testing whether the proportion of locations with snow is lower than the proportion of snow of random locations within the entire nonbreeding home range (95% KDE) for every individual (e.g. by setting a 1/10 ratio presence to random locations).

The proposed analysis will provide an opportunity to assess whether the Rough-legged buzzard selects areas with the lowest snow cover, but will not provide an opportunity to follow the dynamics and will therefore give a misleading overall picture. This is especially true in the spring months. In March-April, Rough-legged buzzards move northeast and are in an area that is not the most open to snow. At this time, areas to the southwest are more open to snow (this can be seen in Figure 3b). If we perform the proposed analysis, the control points for this period would be both to the north (where there is more snow) and to the south (where there is less snow) from the real locations, and the result would be that there is no difference in snow cover.

A step-selection analysis could be used, as we did in our previous work (Curk et al 2020 Sci Rep) with the same Rough-legged buzzards (but during migration, not winter). But this would only give us a qualitative idea, not a quantitative one - that Rough-legged Buzzards move from snow (in the fall) and follow snowmelt progression (in the spring).

At the same time, our analysis gives a complete picture of snow cover dynamics in different parts of the non-breeding range. This allows us to see that if Rough-legged buzzards remained at their fall migration endpoint without moving southwest, they would encounter 14.4% more snow cover (99.5% vs. 85.1%). Although this difference may seem small (14.4%), it holds significance for rodent-hunting birds, distinguishing between complete and patchy snow cover.

Simultaneously, if Rough-legged buzzards immediately flew to the southwest and stayed there throughout winter, they would experience 25.7% less snow cover (57.3% vs. 31.6%). Despite a greater difference than in the first case, it doesn't compel them to adopt this strategy, as it represents the difference between various degrees of landscape openness from snow cover.

Version 3

Abstract

Abstract

Advances in tracking technologies have revealed the diverse migration patterns of birds, which are critical for range mapping and population estimation. While population trends are typically estimated in breeding ranges where birds remain stationary, trends are commonly assessed in non-breeding ranges for species that breed in inaccessible regions such as the Arctic. This task is challenging due to the extensive movements of birds in these ranges, which require a deep understanding of the dynamics of these movements. However, these movements remain understudied, particularly in the mid-latitudes, where many Arctic breeders winter, increasing the risk of misestimating their ranges and numbers. Here, we show that the Arctic breeding raptor Rough-legged buzzard, which overwinters in the mid-latitudes, has a specific wintering strategy: during the entire non-breeding season, they move about 1000 km back and forth under the influence of the advancing snowline, making their non-breeding area dynamic. In essence, this movement represents an extension of the quick migration process, albeit at a slower pace, and we have termed this migration pattern ‘foxtrot migration,’ drawing an analogy to the alternating fast and slow movements of the foxtrot dance. These results highlight the potential errors in range mapping from single mid-winter surveys and emphasize the importance of this migration pattern in assessing the conservation status of bird species. Understanding this migration pattern helps to correctly estimate bird populations in non-breeding areas, which is especially important for species that nest in hard-to-reach regions such as the Arctic.

Introduction

In recent years, advances in tracking technology have greatly improved our understanding of bird migration (Wikelski et al. 2007, Kays et al. 2015, Jetz et al. 2022, Kays and Wikelski 2023). By tracking tagged birds, we can study various migratory behaviors, such as their propensity to migrate, connectivity and philopatry, length of migration routes, variations in flight speed, and movements within a single season (Alerstam et al. 2003, Newton 2008, Alerstam 2011, Berthold et al. 2013, Trierweiler et al. 2013, Schlaich et al. 2023). Different combinations of these behaviors contribute to various bird migration patterns (Chapman et al. 2014, Alerstam and Bäckman 2018, Lislevand et al. 2020). Understanding the migratory patterns of birds is critical to accurately mapping their ranges and estimating population trends. Typically, population trends are assessed in breeding areas, where birds are linked to their nesting sites, allowing reliable estimates of breeding numbers (Reif 2013). However, access to some breeding areas, such as the Arctic, presents logistical and political challenges (Gallo-Cajiao et al. 2023, Koivurova and Shibata 2023). As a result, population trends for these species are often estimated in non-breeding areas where birds aren’t tied to specific locations (Robinson et al. 2005, JIa et al. 2016). This challenges accurately censusing birds and delineating their ranges, especially when birds move extensively in non-breeding ranges during the season, and censuses are conducted only once per non-breeding season. Therefore, understanding bird movements in non-breeding areas is essential for understanding the ecology of birds in these regions and has practical implications for accurately assessing their ranges and trends in their populations.

Many bird species experience significant seasonal changes in food availability within their non-breeding range due to environmental conditions. This often triggers movements across these areas. For Palearctic migrants wintering in Africa, these movements have been studied extensively (Moreau 1972, Trierweiler et al. 2013, Thorup et al. 2017, Schlaich et al. 2023). However, the non-breeding movements of mid-latitude birds remain poorly understood. At the same time, many Arctic birds winter in the mid-latitudes, and understanding their movement patterns could be crucial for interpreting their population trends. In the mid-latitudes, food availability is strongly influenced by environmental factors such as snow cover dynamics. The progressive movement of the snow cover line from north (-east) to south (-west) and back again between October and May in many mid-latitude regions, particularly those exposed to northwesterly wind systems, significantly affects food availability for numerous bird species (Sonerud 1986, Vansteelant et al. 2011). We propose that species dependent on this environmental factor gradually move away from snow-covered areas during winter and then gradually move back in the opposite direction, resulting in a directed and continuous displacement of their non-breeding range.

The Rough-legged buzzard (Buteo lagopus) is an Arctic breeding and mid-latitude wintering raptor (Ferguson-Lees and Christie 2001, Bechard and Swem 2002). Rough-legged buzzards feed mainly on small rodents in the Arctic and its wintering grounds (Tast et al. 2010, Pokrovsky et al. 2014). They prefer open areas for hunting, and trees and tall bushes or uplands for resting. In the Arctic, such areas are the southern and typical tundra (Walker et al. 2005), and in the mid-latitudes, areas with fields and patches of forests (wooded fields). The taiga zone, where there is little open space, is unsuitable for them, although they may nest in the northern taiga zone on the border with the tundra zone (Sundell et al. 2004). Snow cover and day length are important in their life (Terraube et al. 2015, Curk et al. 2020, Pokrovsky et al. 2021). Rough-legged buzzards can only hunt during the day (Pokrovsky et al. 2021), and heavy snow cover makes hunting for small rodents problematic (Sonerud 1986, Vansteelant et al. 2011). These two environmental factors, therefore, affect the availability of prey for Rough-legged buzzards. At the same time, these factors vary considerably in mid-latitudes during winter. Thus, prey availability for Rough-legged buzzards in the mid-latitudes increases until mid-winter if they migrate southwards and after mid-winter if they migrate northwards. We, therefore, assume that Rough-legged buzzards could track prey availability and experience a dynamic non-breeding range.

In this study, we tracked GPS-tagged Rough-legged buzzards for ten years and analyzed their movements in the non-breeding mid-latitude range in relation to snow cover. For this study, we made the following predictions. 1) Rough-legged buzzards would exhibit a directional and seasonal movement pattern during the non-breeding period, moving from the northeast to the southwest and back again. This would result in a dynamic non-breeding range that would continue to move geographically throughout the season. 2) The non-breeding movements would differ from the fall and spring migrations in duration, extent, speed, and direction but would continue throughout the season. Thus, over the entire annual cycle, Rough-legged buzzards would be characterized by a migratory pattern consisting of an alternation of quick (fall and spring migrations) and slow (non-breeding movements) phases. 3) Non-breeding movements would occur in suitable open habitats, while fall and spring migrations would occur in unsuitable forested areas. Thus, vegetation land cover will determine the extent of quick and slow migration phases. 4) During non-breeding movements, Rough-legged buzzards will experience less snow cover than if they had stayed where they arrived at the end of the fall migration. Thus, snow cover dynamics will determine the dynamics of non-breeding movements.

Material and methods

Dataset

For this study, we tracked 43 adult Rough-legged buzzards (35 females and eight males) with the solar GPS-GSM loggers (e-obs GmbH and UKn – University of Konstanz). Twenty-eight birds were fitted with E-obs loggers, 13 birds were fitted with UKn loggers, and two birds were initially fitted with UKn loggers that were later replaced with E-obs loggers. E-obs loggers weighed 45 g, or 3.6% of the bird’s weight, while UKn loggers weighed 15 g, or 1.2% of the bird’s weight. E-obs loggers recorded GPS positions every hour in full battery mode and every 5 hours in normal battery mode. UKn loggers recorded GPS positions every hour with a full battery and every 12 hours with a normal battery. Both models operated continuously 24 hours a day. More detailed information about this dataset can be found in Curk et al. (2022).

The fieldwork was carried out in the Russian Arctic in 2013-2019 at four study sites: Kolguev Island (69°16′N, 48°87′E), Nenetsky Nature Reserve (68°20′N, 53°18′E), Vaigach Island (69°43′N, 60°08′E), and Yamal Peninsula (68°12′N, 68°59′E). For details on capture methods, see Curk et al. (2022); for detailed study descriptions, see Pokrovsky et al. (2015) for Kolguev and Pokrovsky et al. (2019) for Yamal and Nenetsky.

During data pre-processing, we estimated the date of death using an accelerometer and GPS data and excluded tracking data corresponding to the period after the bird’s death. We then removed duplicated timestamps and calculated the mean daily positions of each individual. We partitioned the resulting dataset into several periods: 1) breeding, 2) fall migration, 3)1st part of winter, 4) 2nd part of winter, and 5) spring migration. We estimated the migration dates – the start and stop dates of the spring and fall migrations – using an iterative search procedure for piecewise regression described by Crawley (2007). We created two models for each of the four dates: a simple model and a piecewise one. The simple model was linear, with latitude as the response variable and day of the year as a fixed effect. The piecewise model was also linear, with latitude as the response variable and day of the year as a fixed effect, and included two logical statements in its formula because we expected two linear segments in the fit. Specifically, it used “Day Of the Year < BREAK” to define the left regression (before the estimated date) and “Day Of the Year ≥ BREAK” for the right regression (after the estimated date). We explored different BREAK values to fit the piecewise model, looking for the value that minimized the residual standard error. The BREAK values examined included those associated with the start and end of the fall migration (September 5 to October 5 and October 5 to November 5, respectively), as well as those associated with the start and end of the spring migration (April 10 to May 10 and May 10 to June 10, respectively). We then validated (using the R function ‘anova’) that the piecewise model significantly improved the fit compared to the simple model (p<0.001). We estimated the date between winter’s first and second parts as the day when the mean daily latitude was minimum.

Data analysis

First, we used linear mixed-effects models (R function ‘lmer’ in the library ‘lme4’ (Bates et al. 2015)) to investigate whether or not Rough-legged buzzards migrated during winter. Latitude was the response variable, day of the year was a fixed effect, and individuals and year were included as random effects. Analyses were conducted separately for each migration period (fall, first phase of winter, second phase of winter, and spring). For both phases of the winter migration, we analyzed two additional models with longitude as the response variable instead of latitude. Likelihood ratio tests were used to compare candidate models. The year was not a calendar year but a year between two consecutive breeding seasons. Thus, fall migration, consecutive winter, and consecutive spring have the same value for the year. The day of the year was recalculated consecutively.

Second, we used linear mixed-effects models (R function ‘lmer’ in the library ‘lme4’) to investigate whether migrations’ parameters differ between the migration periods. We analyzed four migration parameters: distance, duration, speed, and direction. The distance was calculated as the distance between two coordinates (start and end of migration) using the R function ‘distm’ in the library ‘geosphere’ (Hijmans 2016). The duration was calculated as the number of days between the start and end of migration. Speed was calculated as the ratio of distance to duration. The direction was calculated as the bearing from the start of the migration coordinates to the end of the migration coordinates using the R function ‘bearing’ in the library ‘geosphere’ (Hijmans 2016). The migration parameter was used as the response variable, the type of migration as a fixed factor, and individuals as a random factor. Likelihood ratio tests were used to compare candidate models. We considered four different parameters of migration (distance, duration, speed, and direction) and four types of migration (fall, first phase of winter, second phase of winter, and spring). The analysis was done separately for each of the migration parameters. Then, we used post hoc comparisons using the R function ‘emmeans’ in the library ‘emmeans’ (Lenth et al. 2019) to compare the estimated means.

In some raptor species, adult females disperse further than males (Mearns and Newton 1984, Serrano et al. 2001, Bildstein 2006, Whitfield et al. 2009). Therefore, we conducted an additional analysis on the effect of sex on migration length using linear mixed-effects models (R function ‘lmer’ in the library ‘lme4’). The migration distance was used as the response variable, sex as a fixed factor, and individuals as a random factor.

Third, we investigated whether vegetation land cover differed between areas crossed during the quick (fall and spring) and slow (winter) migrations. We used the combined Terra and Aqua Moderate Resolution Imaging Spectroradiometer (MODIS) Land Cover Climate Modelling Grid (CMG) (MCD12C1) version 6 dataset (Friedl and Sulla-Menashe 2015). We used a modified Leaf Area Index (LAI) as a classification scheme. We combined all four forest types and savannas into one category (forest) and excluded the categories: water bodies, and unclassified. We, therefore, had five types of vegetation cover: forest, grassland, cropland, shrubland, and urban. We annotated the mean daily positions with the vegetation cover type using the Env-DATA tool (Dodge et al. 2013). We used general linear mixed effects models with a binomial distribution (R function ‘glmer’ in the library ‘lme4’) to investigate whether vegetation cover types differ between migration periods. Presence/absence of the studied vegetation land cover type was used as a response variable, migration type as a fixed factor, and individuals as a random factor. The analysis was done separately for each of the vegetation land cover types.

Fourth, we investigated whether snow cover could drive the slow migration phenomenon. We then compared the snow cover conditions the birds experienced during the winter with two hypothetical snow cover conditions that the birds would have experienced if they had not migrated during the winter. The first hypothetical snow cover condition would have happened if the birds had stayed where they arrived from the north (i.e., where their fall migration ended). To estimate this parameter, we calculated the winter dynamics of the average snow cover at the minimum convex polygons (MCP) occupied by the birds in October and April (northeast of their winter range). A second hypothetical snow cover condition would be if the birds flew immediately to the southwest and spent the whole winter there. To evaluate this, we calculated the winter dynamics of average snow cover on the MCPs occupied by the birds in January and February (southwest of their winter range). We then compared the values obtained for the real snow cover and two hypothetical snow covers using general linear mixed effects models with a binomial distribution (R function ‘glmer’ in the ‘lme4’ library). Presence/absence of snow cover was used as a response variable, type of snow cover (real, 1st hypothetical, or 2nd hypothetical) as a fixed factor, and years as a random factor. The analysis was done separately for each month. We then used post hoc comparisons to compare the estimated means, using the R function ‘emmeans’ in the ‘emmeans’ library (Lenth et al. 2019).

We obtained monthly snow cover data with a spatial resolution of ca 500 meters (Global SnowPack MODIS) from the German Aerospace Center (DLR). This product is based on the Moderate Resolution Imaging Spectroradiometer (MODIS) daily snow cover products MOD10A1 and MYD10A1 (version 6 as provided by the National Snow and Ice Data Center NSIDC), which have been processed to remove the gaps due to cloud cover and polar darkness (Dietz et al. 2015). These processing steps include a combination of data available from different satellites (Aqua and Terra), 3-day temporal moving window filtering, a regional snow line elevation interpolation relying on a Digital Elevation Model (DEM), and a seasonal filter running through the time series for the whole hydrological year (1st of September through August 31st). The proportion of days in which one pixel is snow-covered per month is referred to here as fractional snow cover and is derived from these daily gap-filled rasters. Five MODIS tiles (h19v03, h20v03, h20v04, h21v03 and h21v04) were mosaicked and re-projected to WGS84. Then, for each month from October to April, we calculated 95% minimum convex polygons (MCPs) for the distribution of Rough-legged buzzards using the R function ‘mcp’ in library ‘adehabitatHR’ (Calenge 2006). We extracted mean snow cover values from each MCP from every monthly snow cover raster separately, using the R library ‘raster’ (Hijmans and van Etten 2023).

All calculations were performed using R version 4.2.2 ‘Innocent and Trusting’ (R Development Core Team 2022) and RStudio version 353 ‘Elsbeth Geranium’ (Posit team 2022).

Results

Migration of Rough-legged buzzards – always on the move

Except during the breeding season, Rough-legged buzzard migration continues throughout the year, even after the birds’ arrival at their traditionally recognized ‘wintering grounds’ (Figure 1a, b). Rough-legged buzzards started their fall migration on 28 September (hereafter mean±sd for the day of the year: 271±11, n=31) and ended on 12 October (285±11, n=33). The mean latitude/longitude where the birds ended their fall migration was 55.57±1.92°/49.35±5.63° (Figure 1a). During the winter, birds continued to migrate at a slower pace down to 49.53±2.01° latitude (on 5 February, 36±40, n=23) and 34.29±5.11° longitude (on 24 January, 24±47, n=23). Afterward, during the second part of the winter, birds returned to 55.52±2.63° latitude and 49.79±8.24° longitude to start the spring migration (Figure 1a). Rough-legged buzzards started their spring migration to the Arctic on 27 April (117±7, n=27) and arrived at the breeding grounds on 15 May (135±8, n=18).

Migration of Rough-legged buzzards.

Q – quick phase. Qf – Quick fall phase (orange), Qs – Quick spring phase (yellow), S1 – Slow phase, 1st part (light blue), S2 – Slow phase, 2nd part (dark blue). a) Change in the latitude of 43 Rough-legged buzzards during the year, red line – mean latitude of all birds, black vertical lines – mean dates of start and end of the migration phases, blue vertical line – mean date of the minimum latitude. Grey, sky blue, and piggy pink shaded areas – standard deviation of the means. b) Migration map. c) Difference in the migration parameters between the migration phases. Lines on the direction plot (down, right) represent the mean value for each bird; arrows represent the mean direction for each phase. Boxes on the boxplots show the interquartile range, the whiskers – maximum and minimum values.

In the following, we will refer to the spring and fall migrations as the quick phase, the non-breeding movement to the lowest point of latitude as the 1st part of the slow phase, and the movement from the lowest point of latitude to the starting point of the spring migration as the 2nd part of the slow phase (Figure 1a). For both quick and slow phases of the migration, linear mixed-effects models with the season as a fixed factor received higher support from the likelihood ratio test (p < 0.001, Tables S1-S3).

Quick-slow phase features comparison

During the quick phase, individual birds flew greater distances in a shorter time, i.e., at a faster rate, than during the slow phase. After arriving at what is traditionally known as the wintering grounds, the direction of migration changed, so the direction of quick and slow phases also differed (Table 1, Figure 1c). The quick phase was 1415±50 km long (hereafter mean±sd), whereas the slow phase (one part) was 1026±55 km, i.e., 389±60 km shorter (p<0.001, Table S4, Figure 1c). During the quick phase, birds flew for 15±3 days, and one part of the slow phase lasted 100±4 days, i.e., 85±5 days longer (p<0.001, Table S5, Figure 1c). At the same time, the second part of the slow phase was 54±7 days shorter than the first (p<0.001, Table S5, Figure 1c). The migration speed was 104±6 km/day during the quick phase and 12±7 km/day during the slow phase, i.e., about eight times higher (p<0.001, Table S6, Figure 1c). During the fall migration, birds moved in the SSW direction (7±2 deg), then turned 50±3 deg (p<0.001, Table S7, Figure 1c) to the west and started their 1st slow phase until mid-winter. After that, they turned back to the NEE direction (57±2 deg) and performed their 2nd slow phase for several months until they turned 54±3 deg (p<0.001, Table S7, Figure 1c) to the north and started their spring migration. As a result of additional analysis of the effect of sex on migration length, we found no significant difference between the migration distances of males and females (Table S8).

Vegetation land cover during migration

During quick phases, Rough-legged buzzards cross the forest zone, while during the slow phase, they migrate within the grassland and cropland zone (Figure 2). Rough-legged buzzards migrated fast across the tundra zone on the north in the Arctic and then through the taiga zone. Therefore, during the quick phase, the three most common vegetation land cover types were forest (44.5±2.9 %, hereafter, percentage of all mean daily positions annotated with the given vegetation type ± sd), shrublands (29.9±3 %), and grasslands (24.7±2.8 %, Figure 2a). During the slow phase, the three most common vegetation land cover types were grasslands (65.1±6.2 %), croplands (26.9±6.3 %), and forests (4.9±1.4 %, Figure 2b). According to the linear mixed-effects models, the percentage of all vegetation land cover types differed between the slow and quick phases (p<0.001, Table S9), except for the urban lands. Urban lands were more common during the slow than quick phase (Figure 2). However, this type has been annotated for too few birds to make an adequate comparison.

Vegetation land cover during quick and slow phases of the migration.

a) Quick phase (spring and fall periods together). b) Slow phase (1st and 2nd parts together). On both a) and b), the bars show the percentage of all mean daily positions annotated with the given vegetation type ± sd. c) Migration map.

Snow cover – the main reason for the dynamic winter range

During the slow phase of the migration, Rough-legged buzzards experienced snow cover ranging from 4.8±1.0% in October (hereafter mean±sd) to 85.2±4.6% in February (Figure 3). If birds spent the winter in the place where they arrived after the fall migration, they would experience snow cover conditions ranging from 4.6±0.6% in October to 99.5±0.1% in February (Figure 3b, green line). If birds fly directly to the southeast and stay there for the whole winter, they would experience snow cover conditions ranging from 1.4±0.2 % in October to 81.1±5.0 % in January (Figure 3b, red line). Thus, if birds fly immediately to the southwest and stay there until the end of the winter, they will find conditions with less snow cover in spring (p<0.001, Table S10). And if birds stay where they ended the fall migration, they will find themselves in situations with more snow cover (p<0.001, Table S10). In the latter case, the difference between real and hypothetical situations is not as pronounced (85.2% vs. 99.5%), but more means that snow cover will be close to 100% for several months in this hypothetical situation (Figure 3b).

Snow cover conditions during the slow phase of the migration.

a) 95% Minimum convex polygons (MCPs) of Rough-legged buzzards during winter. Arrows indicate the direction of the movement across months. OCT – October, NOV – November, DEC – December, JAN – January, FEB – February, MAR – March, APR – April. b) Snow cover conditions for the real situation (black) and two hypothetical situations – if birds spend the winter in the place where they arrived after the fall migration (green) and if birds fly directly to the southwest and stay there all winter (red). Dots represent mean values, error lines – standard deviations.

Discussion

Our study identified and characterized a bird migration pattern consisting of an alternation of quick (fall and spring migrations) and slow (non-breeding movements) phases. This migration pattern causes the non-breeding range of birds to shift and become dynamic throughout the season. This has obvious implications for range delineation and assessment of bird population trends.

Non-breeding movements and migratory patterns

Different bird species employ various wintering strategies, leading to diverse migration patterns. One strategy is residency, where birds arriving at non-breeding grounds remain there throughout the season, undertaking only minor foraging flights (Figure 4a). Their home ranges in non-breeding areas are compatible with those in breeding areas (Kjellén et al. 1997, Alerstam et al. 2006, Newton 2008). Another strategy, initially identified by Moreau (Moreau 1972), is itinerancy, wherein birds fly between several sites separated by distances ranging from ten to several hundred kilometers throughout the non-breeding period, spending 1-2 months at each site (Figure 4b). Most Palearctic-African species adopt this itinerant behavior (Trierweiler et al. 2013, Thorup et al. 2017, Schlaich et al. 2023). The third strategy investigated here involves birds moving slowly and steadily throughout the non-breeding season, initially in one direction and then in the opposite direction. In 2009, Strandberg et al. used GPS transmitters to track Common buzzards (Buteo buteo). Despite some transmitters in their study ceasing to function mid-winter, the authors observed a phenomenon they termed “prolonged autumn migration,” wherein birds arriving at their wintering grounds did not halt migration but continued moving at a slower pace (Strandberg et al. 2009). In our study, we observed this phenomenon throughout the annual cycle. We demonstrated that, unlike itinerancy, movements in the non-breeding area are essentially a continuation of migration at a slower pace, influenced by external factors such as snow cover dynamics. Considering this behavior as an extension of migration and examining the complete life cycle of such species, it involves a quick phase during the transition between breeding and non-breeding areas, followed by a slow phase of directed and seasonal movement within the non-breeding area, and finally, another quick phase towards the breeding area. Drawing an analogy to the alternating quick and slow movements of the foxtrot dance, we propose the term ‘foxtrot migration’ for this seasonal movement to offer a concise and easily understandable description (Figure 4c). Consequently, we suggest referring to the non-breeding range of species exhibiting itinerancy or foxtrot migration as the ‘dynamic range’ (Figure 4b, c).

Winter strategies and non-breeding range scheme.

a) Residency, b) Itinerancy, c) Foxtrot migration. The color bar along the y-axis corresponds to the color coding of the habitats of the chart on the right.

Debates may arise regarding introducing a new term for this phenomenon and whether it should fall under the general term “nomadism” or be categorized as a modified form of another migratory pattern, such as “slow directional itinerancy.” We assert that behavioral pattern names should aid in understanding movement strategies and avoid confusion. One might ask whether it is necessary to distinguish between migration and wintering at all, given that migration is essentially a matter of tracking spatial and temporal changes in favorable conditions. Authors studying Montagu’s harriers (Circus pygargus) argue that such generalization hampers understanding of wintering strategies (Schlaich et al. 2023). Instead, Schleich et al. (2023) propose explicitly considering the functions of animal-used sites, particularly their contributions to “wintering” and “refueling for migration.” For instance, cuckoos (Canorus sp.) utilize stopovers for both peak foraging conditions and refueling for long flights to subsequent sites (Thorup et al. 2017). Conversely, Montagu’s harriers exhibit negligible refueling between sites due to short distances and energy-efficient soaring flight (Trierweiler et al. 2013, Schlaich et al. 2023). In our case, Rough-legged buzzards do not make several long stopovers, unlike species experiencing itinerancy, but move gradually in one direction and then return during the whole season (Figure 1a). As Schleich et al. (2023) suggested, examining the functions of animal-used sites reveals another difference between these strategies. The difference is that itinerant birds wait for better conditions at subsequent sites. In contrast, in our case, the southeastern areas that birds reach after several months are suitable and offer better conditions than the northeastern areas, as they have less snow throughout the whole non-breeding season (Figure 3). All these differences highlight the discrepancy between the wintering strategies referred to as itinerancy and those described in this study. Therefore, using another term - foxtrot migration - for this phenomenon is more appropriate.

The dynamic of the foxtrot migration

Our study affirmed the presence of foxtrot migration and a dynamic non-breeding range in Rough-legged buzzards. The taiga zone posed as an unfavorable habitat during the quick phase of foxtrot migration (Figure 2), given the difficulty of Rough-legged buzzards locating open areas for hunting in this habitat. Conversely, the grassland and cropland zone served as favorable habitats during the slow phase throughout the entire non-breeding period (Figure 2), offering numerous open areas for hunting. Snow cover was the external factor driving their continual 1000 km southwest movement during winter (Figure 3). Our analysis revealed that if Rough-legged buzzards remained at their fall migration endpoint without moving southwest, they would encounter 14.4% more snow cover (99.5% vs. 85.1%, Figure 3). Although this difference may seem small (14.4%), it holds significance for rodent-hunting birds, distinguishing between complete and patchy snow cover. Simultaneously, if Rough-legged buzzards immediately flew to the southwest and stayed there throughout winter, they would experience 25.7% less snow cover (57.3% vs. 31.6%, Figure 3b). Despite a greater difference than in the first case, it doesn’t compel them to adopt this strategy, as it represents the difference between various degrees of landscape openness from snow cover. However, it is noteworthy that one bird followed this strategy by flying directly to the southwest and remaining there throughout the winter (Figure S1), exhibiting a residency strategy. This observation suggests that the distinction between winter strategies is flexible.

African-Palearctic migrants do not strictly adhere to a binary division between residency and itinerancy but demonstrate a continuum between these two strategies. Many species in Africa exhibit a mixture of both strategies in varying proportions; see Schlaich et al. (2023) and reference therein. Similarly, in the mid-latitudes, species may not only show any of the three strategies but also a considerable degree of individual and regional variation. For example, Rough-legged buzzards from western Arctic regions that overwinter in mid-latitude areas where snow cover dynamics are less pronounced than in eastern regions are likely to have a higher proportion of birds with a resident strategy. At the same time, Rough-legged buzzards in North America also exhibit a foxtrot migration pattern, although with variations. There, birds make a direct and rapid migration, usually across the boreal forest, followed by slower, more facultative movements once birds get south of the boreal. However, there is a lot of variation among individuals and even regions; there are more non-breeding slow movements in eastern North America but less in western North America (Neil Paprocki pers. comm.). Thus, Rough-legged buzzards in North America lie along a gradient between foxtrot migration and residency, likely influenced by different snow cover dynamics on the east and west coasts.

The foxtrot migration pattern is expected to be observed in many migratory species that exhibit distinct seasonal cycles in their non-breeding range. This migration pattern is expected to be prevalent for species living in regions where snow cover is a significant determinant of food availability, which is the case in large parts of the mid-latitudes. Therefore, understanding bird migration patterns is critical to accurately mapping ranges and assessing population trends in this region.

Mapping ranges and assessing population trends

The implementation of our study is twofold: 1) the use of mid-winter bird surveys to determine non-breeding range may yield inaccurate results for species with dynamic range (Figure 5a), and declines in abundance within a particular segment of the non-breeding range may indicate changes in range dynamics rather than widespread declines in species abundance (Figure. 5b).

Mapping ranges and assessing population trends for the species with the dynamic non-breeding ranges.

a) Mid-winter counts, often used to determine winter range, may give a misleading representation of the non-breeding range for species with dynamic range. b) Changing the conservation status of a species based on reduced abundance in a particular area may not be appropriate for species with dynamic range. Red – breeding range, blue – winter range.

In North America and Europe, the number of non-breeding birds is typically estimated once a season in mid-winter. In North America, these estimates are made during the Christmas counts, usually once in late December (National Audubon Society 2024), while in Europe, they are made during the IWC counts, usually once in January or February (Wetlands International 2024). This approach can lead to errors in range estimation for birds with dynamic ranges. For such species, habitats occupied by birds in fall and spring will be listed as “uncommon” at best, while habitats occupied by birds only in mid-winter will be listed as “common.” However, the situation is the opposite if we consider the time birds spend in these habitats (Figure 5a). For example, based on our study of the Rough-legged buzzard, during mid-winter, the species is predominantly present in the southwestern portion of its non-breeding range, with only a tiny proportion present in the rest of the region. As a result, a map of the non-breeding range may show the species as “common” in the southwest and “uncommon” in the northeast. This map would be inaccurate because, during the entire non-breeding period, Rough-legged buzzards spend both fall and spring in the northeastern part and only mid-winter in the southwestern part.

To address this, continuous year-round GPS tracking of the species provides a means to track bird locations throughout the non-breeding season, facilitating the creation of accurate distribution maps, particularly for species that exhibit dynamic ranges. We advocate representing temporal heterogeneity (range dynamics) on maps as distinct zones, denoting periods when the species is abundant in a given area. To distinguish temporal from spatial heterogeneity, we recommend using lines to delineate the boundaries of these zones, rather than color shading, and incorporating numbers to denote the months of species abundance in a given zone (Figure 5a). We suggest ecologists include dynamic non-breeding ranges in descriptions and range maps for foxtrot migratory and itinerant bird species.

Population counts for a species are often limited to a portion of its range. Therefore, conclusions about conservation status drawn from such counts may be misleading. A decline in abundance within a particular portion of the non-breeding range may indicate changes in range dynamics rather than a general decline in the species (Figure 5b). For example, climate change may affect snow cover dynamics, reducing its intensity in northern regions. As a result, species whose range dynamics depend on snow cover may choose to remain in the northern areas and not migrate as far south as they traditionally have. Despite this shift in range dynamics, overall species abundance may remain unchanged. Similar patterns have affected Rough-legged buzzards in some areas of the European non-breeding range.

A 2022 Dutch study found a decline in wintering Rough-legged buzzards over the last 40 years (Hornman et al. 2022). On the one hand, this may represent a conservation concern. On the other hand, applying the rationale of the dynamic range, the apparent local decline may be attributed to climate change, resulting in less comprehensive snow coverage in the northeastern wintering areas of Rough-legged buzzards relative to the Netherlands. Such a shift in snow coverage makes it less probable for the birds to migrate to the Netherlands for overwintering. This proposition is further supported by a study of the winter population dynamics of Rough-legged buzzards in the Netherlands in 2011, showing that the main winter population peak occurred in late December, with many birds migrating (Vansteelant et al. 2011). Vansteelant et al. (2011) also found that the main migration occurred after heavy snowfall in northern Europe, supporting our foxtrot migration explanation for this decline. Therefore, investigating the dynamic range is critical to understanding a species’ range and effectively assessing its conservation status.

Conclusion

Our study sheds light on the overwintering strategy of the Rough-legged buzzard, an Arctic breeder that migrates to mid-latitudes during the non-breeding season. We have observed a migration pattern in which these birds travel about 1000 km round trip in response to changing snowline conditions, which we call the ‘foxtrot migration’ for its similarity to the alternating quick and slow movements of the foxtrot dance. Our results highlight the importance of this migration pattern and the concept of ‘dynamic range’ in revealing potential inaccuracies in range mapping from single midwinter surveys and suggest that species distribution maps should represent temporal variations in range dynamics as separate zones. Understanding and accounting for this migratory behavior is critical for accurately assessing bird populations in non-breeding areas. This is especially important for species such as Arctic nesting birds, where population assessments are primarily conducted in non-breeding ranges due to the logistical and political challenges of the Arctic.

Supporting information